---
title: "Reading and Cleaning Data with {mnirs}"
vignette: >
    %\VignetteIndexEntry{Reading and Cleaning Data with {mnirs}}
    %\VignetteEngine{quarto::html}
    %\VignetteEncoding{UTF-8}
format:
    html:
        embed-resources: true
        toc: true
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
        dpi: 300
        out.width: "100%"
        fig.align: "centre"
        eval: requireNamespace("ggplot2", quietly = TRUE)
---

# Introduction

With modern wearable **muscle near-infrared spectroscopy (mNIRS)** devices it is growing ever easier to collect local muscle oxygenation data during dynamic activities. The real challenge comes with deciding how to clean, filter, process, and eventually interpret those data.

The `{mnirs}` package aims to provide standardised, reproducible methods for processing and analysing NIRS data, helping practitioners detect meaningful signals from noise and improve our confidence in interpreting information and applying that information to the clients we work with.

In this vignette we will demonstrate how to:

* 📂 Read data files exported from commercially available wearable NIRS devices, and import NIRS channels into a standard data frame format with metadata, ready for further processing.

* 📊 Plot and visualise data frames of class *`"mnirs"`*.

* 🔍 Retrieve metadata stored with data frames of class *`"mnirs"`* to avoid repetitively specifying which channels to process.

* 🧹 Detect and replace local outliers, invalid values, and interpolate across missing data.

* ⏱️ Resample data to a higher or lower sample rate, to correct irregular sampling periods or match the frequency of other data sources.

* 📈️ Apply digital filtering to optimise signal-to-noise ratio for the responses observed in our data.

* ⚖️ Shift and rescale across multiple NIRS channels, to normalise signal dynamic range while preserving absolute or relative scaling between muscle sites and NIRS channels.

* 🔀 Demonstrate a complete data wrangling process using pipe-friendly functions.

::: {.callout-note}
*`mnirs`* is currently [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental). Functionality may change! Stay updated on development and follow releases at [github.com/jemarnold/mnirs](https://github.com/jemarnold/mnirs).

*{mnirs}* is designed to process NIRS data, but it can be used to read, clean, and process other time series datasets which require many of the same processing steps. Enjoy!
:::

# 📂 Read data from file

We will read an example data file with two NIRS channels from an incremental ramp cycling assessment recorded with *Moxy* muscle oxygen monitor.

First, install and load the *`mnirs`* package and other required libraries.

*`mnirs`* can be installed with `remotes::install_github("jemarnold/mnirs")`.

```{r}
#| label: setup
#| warning: false

# remotes::install_github("jemarnold/mnirs") ## install development version
library(ggplot2) ## load for plotting
library(mnirs) 
```

The first function to call will almost always be `read_mnirs()`. This is used to read data from *.csv* or *.xls(x)* files exported from common wearable NIRS devices. It will extract and return the data table and metadata for further processing and analysis.

See `read_mnirs()` for more details.


## `read_mnirs()`

* `file_path`

    Specify the location of the data file, including file extension. e.g. `"./my_file.xlsx"` or `"C:/myfolder/my_file.csv"`.

::: {.callout-note}
### Example data files

A few example data files are included in the *`mnirs`* package. File paths can be accessed with `example_mnirs()`
:::

* `nirs_channels`
    
    At least one *NIRS* channel name must be specified from the data table in the file. Multiple channel names can be entered as a vector.

* `time_channel`

    A *time* or *sample* channel name from the data table can be specified. If left blank, the function will attempt to identify the time column automatically, however the best practice is to specify the time_channel column explicitly.

* `event_channel`

    Optionally, A channel can be specified which indicates *events* or *laps* in the data table.\
\
These channel names are used to detect the data table within the file, and must match exactly with text strings in the file on the same row. We can rename these channels when reading data by specifying a named character vector: 

```{r}
nirs_channels = c(new_name1 = "original_name1", 
                  new_name2 = "original_name2")
```

* `sample_rate`

    The sample rate of the exported data file (in Hz) can either be specified explicitly, or it will be estimated from `time_channel`.\
\
Automatic detection usually works well unless there is irregular sampling, or the `time_channel` is a count of samples rather than a time value. For example, *Oxysoft* exports a sample number rather than time values. In this specific case, the function will recognise and read the correct sample rate from the file metadata. However, in most cases sample_rate should be defined explicitly if known.

* `add_timestamp`

    If the `time_channel` is detected as date-time format (e.g.; *hh:mm:ss*), by default it will be converted to numeric time in seconds. If specified, this option will preserve a seperate timestamp column of those date-time values.

* `zero_time`

    If `time_channel` values start at a non-zero value, this option will re-calculate time starting from zero. If `time_channel` was converted from date-time format, it will always be re-calculated from zero regardless of this option.

* `keep_all`

    By default, only the channels explicitly specified above will be returned in a data frame. This option will return all columns detected from the data table in the file. Blank/empty columns will be omitted and duplicated or invalid column names will be repaired. Columns should be checked to confirm correct naming if duplicates are present.

* `verbose`

    This and many other *`mnirs`* functions may return warnings and informational messages which are useful for troubleshooting and data validation. This option can be used to silence those messages. *`mnirs`* messages can be silenced globally for a session by setting `options(mnirs.verbose = FALSE)`.

```{r}
## {mnirs} includes sample files from a few NIRS devices
example_mnirs()

## partial matching will error if matches multiple
try(example_mnirs("moxy"))

## call an example mNIRS data file
file_path <- example_mnirs("moxy_ramp") 

data_table <- read_mnirs(
    file_path,
    nirs_channels = c(
        smo2_right = "SmO2 Live",        ## identify and rename channels
        smo2_left = "SmO2 Live(2)"
    ),
    time_channel = c(time = "hh:mm:ss"), ## date-time format will be converted to numeric
    event_channel = NULL,                ## left blank, not currently used in analysis
    sample_rate = NULL,                  ## if blank, sample_rate will be estimated from time_channel
    add_timestamp = FALSE,               ## omit the date-time timestamp column
    zero_time = TRUE,                    ## recalculate time values from zero
    keep_all = FALSE,                    ## return only the specified data channels
    verbose = TRUE                       ## show warnings & messages
)

## Note the above info message that sample_rate was estimated correctly at 2 Hz ☝
## ignore the warnings about irregular sampling for now, we will resample later

data_table
```


# 📊 Plot *`mnirs`* data

*`mnirs`* data can be easily viewed by calling `plot()`. This generic plot function uses `{ggplot2}` and will work on data frames generated or read by *`mnirs`* functions where the metadata contain `class = *"mnirs"*`.

```{r}
## note the hidden plot option to display time values as `h:mm:ss`
plot(data_table, label_time = TRUE)
```

*`mnirs`* includes a custom `*ggplot2*` theme and colour palette available with `theme_mnirs()` and `palette_mnirs()`. See those documentation references for more details.


# 🔍 Metadata stored in *`mnirs`* data frames

Data frames generated or read by *`mnirs`* functions will return `class = *"mnirs"*` and contain metadata, which can be retrieved with `attributes(data)`. 

Instead of re-defining values like our channel names or sample rate, certain *`mnirs`* functions can automatically retrieve them from metadata. They can always be overwritten manually in subsequent functions, or by using a helper function `create_mnirs_data()`.

See `create_mnirs_data()` for more details about metadata.

```{r}
## view metadata (omitting item two, a list of row numbers)
attributes(data_table)[-2]
```


# 🧹 Replace local outliers, invalid values, and missing values

We can see some data issues in the plot above, so let's clean those with a single wrangling function `replace_mnirs()`, to prepare our data for digital filtering and smoothing.

*`mnirs`* tries to include basic functions which work on vector data, and convenience wrappers which combine functionality and can be used on multiple channels in a data frame at once. Let's explain the functionality of this data-wide function, and for more details about the vector-specific functions see `replace_mnirs()`.


## `replace_mnirs()`

* `data`

    Data-wide functions take in a data frame, apply processing to all channels specified, then return the processed data frame. *`mnirs`* metadata will be passed to and from this function.\
\
*`mnirs`* functions are also pipe-friendly for Base R 4.1+ (`|>`) or `{magrittr}` (`%>%`) pipes to chain operations together (see below).

* `nirs_channels`

    Specify which column names in `data` will be processed, i.e. the response variables. If not specified, these channels will be retrieved from *`mnirs`* metadata. Channels in the data but not explicitly specified will be passed through unprocessed to the returned data frame.

* `time_channels`

    The time channel can be specified, i.e. the predictor variable, or retrieved from *`mnirs`* metadata.

* `invalid_values`, `invalid_above`, or `invalid_below`

    Specific invalid values can be replaced, e.g. if a NIRS device exports `0`, `100`, or some other fixed value when signal recording is in error. If spikes or drops are present in the data, these can be replaced by specifying values above or below which to consider invalid. If left as `NULL`, no values will be replaced.

* `outlier_cutoff`

    Local outliers can be detected using a cutoff calculated from the local median value. A default value of `3` is recommended (i.e., ± 3 SD about the local median). If left as `NULL`, no outliers will be replaced.

* `width` or `span`

    Local outlier detection and median interpolation use a rolling local window specified by one of either `width` or `span`. `width` defines a number of samples centred around the local index being evaluated (`idx`), whereas `span` defines a range of time in units of `time_channel`.

* `method`

    Missing data (`NA`), invalid values, and local outliers specified above can be replaced via interpolation or fill methods; either `"linear"` interpolation, fill with local `"median"`, or `"locf"` (*"last observation carried forward"*).\
\
`NA`s can be passed through to the returned data frame with `method = "none"`. However, subsequent processing & analysis steps may return errors when `NA`s are present. Therefore, it is good practice to deal with missing data early during data processing.

* `verbose`

    As above, an option to toggle warnings and info messages.

```{r}
data_cleaned <- replace_mnirs(
    data_table,         ## channels will be retrieved from metadata
    invalid_values = 0, ## known invalid values in the data
    invalid_above = 90, ## remove data spikes
    outlier_cutoff = 3, ## recommended default value
    width = 10,         ## local window to detect local outliers and replace missing values
    method = "linear",  ## linear interpolation over `NA`s
    verbose = TRUE
)

plot(data_cleaned, label_time = TRUE)
```

That cleaned up all the obvious data issues.


# ⏱️ Resample data

Say we have NIRS data recorded at 25 Hz, but we are only interested in exercise responses over a time span of 5-minutes, and our other outcome measure heart rate data are only recorded at 1 Hz anyway. It may be easier and faster to work with our NIRS data down-sampled from 25 to 1 Hz.

Alternatively, if we have something like high-frequency EMG data, we may want to up-sample our NIRS data to match samples for easier synchronisation and analysis (*although, we should be cautious with up-sampling as this can artificially inflate statistical confidence with subsequent analysis or modelling methods*).


## `resample_mnirs()`

* `data`

    This function takes in a data frame, applies processing to all channels specified, then returns the processed data frame. *`mnirs`* metadata will be passed to and from this function.

* `time_channel` & `sample_rate`

    If the data contain *`mnirs`* metadata, these channels will be detected automatically. Or they can be specified explicitly.

* `resample_rate`

    Resampling is specified as the desired number of samples per second (Hz). The default `resample_rate` will resample back to the existing `sample_rate` of the data. This can be useful to accomodate for irregular sampling with unequal time values. Linear interpolation is used to resample `time_channel` to round values of the `sample_rate`.

```{r}
data_resampled <- resample_mnirs(
    data_cleaned,      ## channels retrieved from metadata
    resample_rate = 2, ## or leave blank as default `resample_rate = sample_rate`
    method = "linear", ## linear interpolation across any new samples
    verbose = TRUE     ## will confirm the output sample rate
)

## note the altered "time" values from the original data frame 👇
data_resampled
```


::: {.callout-note}
## Use `resample_mnirs()` to smooth over irregular or skipped samples

If we see a warning from `read_mnirs()` about duplicated or irregular samples like we saw above, we can use `resample_mnirs()` to restore `time_channel` to a regular sample rate and interpolate across skipped samples.
:::


# 📈 Digital filtering

To improve the signal-to-noise ratio in our dataset without losing information, we should apply digital filtering to smooth the data.



## Choosing a digital filter

There are a few digital filtering methods available in *`mnirs`*. Which option is best for *you* will depend in large part on the sample rate of the data and the frequency of the response/phenomena being observed.

Choosing filter parameters is an important processing step to improve signal-to-noise ratio and enhance our subsequent interpretations. Over-filtering the data can introduce data artefacts which can negatively influence the signal analysis and interpretations just as much as trying to analyse overly-noisy raw data.

It is perfectly valid to choose a digital filter by iteratively testing filter parameters until the signal or response of interest appears to be visually optimised with minimal data artefacts, to your satisfaction.

We will discuss the process of choosing a digital filter more in depth in another vignette `<currently under development>`.


## `filter_mnirs()`

* `data`

    This function takes in a data frame, applies processing to all channels specified, then returns the processed data frame. *`mnirs`* metadata will be passed to and from this function.

* `nirs_channels`, `time_channel`, & `sample_rate`

    If the data contain *`mnirs`* metadata, these channels will be detected automatically. Or they can be specified explicitly.

* `na.rm`

    This important argument is left as `FALSE` by default. This function will return an error if any missing data (`NA`) are detected in the response variables (`nirs_channels`). Setting `na.rm = TRUE` will bypass these `NA`s and preserve them in the returned data frame, but this must be opted into explicitly in this function and elsewhere.


### Smoothing-spline

* `method = "smooth_spline"`

    The default non-parametric cubic smoothing spline is often a good first filtering option when exploring the data, and works well over longer time spans. For more rapid and repeated responses, a smoothing-spline may not work as well.

* `spar`

    The smoothing parameter of the cubic spline will be determined automatically by default, or can be specified explicitly. See `stats::smooth.spline()`


### Butterworth digital filter

* `method = "butterworth"`

    A Butterworth low-pass digital filter (specified by `type = "low"`) is probably the most common method used in mNIRS research (whether appropriately or not). For certain applications such as identifying a signal with a known frequency (e.g. cycling/running cadence or heart rate), a pass-band or a different filter type may be better suited.

* `type`

    The filter type is specified as either `c("low", "high", "stop", "pass")`.

* `order`

    The filter order number, specifying the number of passes the filter performs over the data. The default `order = 2` will often noticably improve the filter over a single pass, however higher orders above ~`4` can begin to introduce artefacts, particularly at sharp transition points.

* `W` or `fc`

    The cutoff frequency can be specified either as `W`; a fraction (between `[0, 1]`) of the [*Nyquist frequency*](https://www.youtube.com/watch?v=IZJQXlbm2dU), which is equal to half of the `sample_rate` of the data. Or as `fc`; the cutoff frequency in Hz, where this absolute frequency should still be between 0 Hz and the Nyquist frequency.

For filtering vector data and more details about Butterworth filter parameters, see `filter_butter()`.


### Moving average

* `method = "moving_average"`

    The simplest smoothing method is a moving average filter applied over a specified number of samples or timespan. Commonly, this might be a 5- or 15-second centred moving average.

* `width` or `span`

    Moving average filtering occurs within a rolling local window specified by one of either `width` or `span`. `width` defines a number of samples centred around the local index being evaluated (`idx`), whereas `span` defines a range of time in units of `time_channel`.

For filtering vector data and more details about the moving average filter, see `filter_moving_average()`


## Apply the filter

Let's try a *Butterworth low-pass* filter, and we'll specify some empirically chosen filter parameters for these data. See `filter_mnirs()` for further details on each of these filtering methods and their respective parameters.


```{r}
data_filtered <- filter_mnirs(
    data_resampled,         ## channels retrieved from metadata
    method = "butterworth", ## Butterworth digital filter is a common choice
    type = "low",           ## specify a low-pass filter
    order = 2,              ## filter order number
    W = 0.02,               ## filter fractional critical frequency
    na.rm = TRUE            ## explicitly preserve any NAs
)

## we will add the non-filtered data back to the plot to compare
plot(data_filtered, label_time = TRUE) +
    geom_line(
        data = data_cleaned, 
        aes(y = smo2_left, colour = "smo2_left"), alpha = 0.4
    ) +
    geom_line(
        data = data_cleaned, 
        aes(y = smo2_right, colour = "smo2_right"), alpha = 0.4
    )
```


That did a nice job reducing the high-frequency noise in our data, while preserving the sharp edges, such as the rapid reoxygenation after maximal exercise.

# ⚖️ Shift and rescale data

NIRS values are not measured on an absolute scale (even, arguably percent (%) saturation). Therefore, we may need to adjust or calibrate our data to normalise NIRS signal values between muscle sites, individuals, trials, etc. depending on our intended comparison.

For example, we may want to set our mean baseline value to zero for all NIRS signals at the start of a recording. Or we may want to compare signal kinetics (the rate of change or time course of a response) after rescaling signal amplitudes to a common dynamic range.

These functions allow us to either shift NIRS values up or down while preserving the dynamic range (the absolute amplitude from minimum to maximum values) of our NIRS channels, or rescale the data to a new dynamic range with larger or smaller amplitude.

We can group NIRS channels together to preserve the absolute and relative scaling among certain channels, and modify that scaling between other groups of channels.



## `shift_mnirs()`

* `data`

    This function takes in a data frame, applies processing to all channels specified, then returns the processed data frame. *`mnirs`* metadata will be passed to and from this function.

* `nirs_channels`

    In these functions, channels should be grouped by providing a list (e.g. `list(c(A, B), c(C))`) where each group will be shifted to a common scale, and separate scales between groups. The relative scaling between channels will be preserved within each group, but lost between groups.\
\
`nirs_channels` should be specified explicitly to ensure the intended grouping structure is returned. The default *`mnirs`* metadata will group all NIRS channels together.

* `time_channel`

    If the data contain *`mnirs`* metadata, this channels will be detected automatically. Or it can be specified explicitly

* `to` or `by`

    The shift amplitude can be specified by either shifting signals `to` a new value, or shifting signals `by` a fixed amplitude, given in units of the NIRS signals.

* `width` or `span`

    Shifting can be performed on the mean value within a window specified by one of either `width` or `span`. `width` defines a number of samples centred around the local index being evaluated (`idx`), whereas `span` defines a range of time in units of `time_channel`.\
\
e.g. `width = 1` will shift from the single minimum sample, whereas `span = 1` would shift from the minimum mean value of all samples within one second.

* `position`

    Specifies how we want to shift the data; either shifting the *"min"*, *"max"*, or *"first"* sample(s).


For this data set, we want to shift each NIRS channel so that the mean of the 2-minute baseline is equal to zero, which would then give us a change in deoxygenation from baseline during the incremental cycling protocol.

::: {.callout-note}
## `{tidyverse}`-style channel name specification

This is a good time to note that here and in most `*mnirs*` functions, data channels can be specified using `{tidyverse}`-style naming; Data frame column names can be specified either with quotes as a character string (`"smo2"`), or as a direct symbol (`smo2`).

`{tidyselect}` support functions (e.g. `starts_with()`, `matches()`) can also be used.
:::

```{r}
data_shifted <- shift_mnirs(
    data_filtered,     ## un-grouped nirs channels to shift separately 
    nirs_channels = list(smo2_left, smo2_right), 
    to = 0,            ## NIRS values will be shifted to zero
    span = 120,        ## shift the first 120 sec of data to zero
    position = "first"
)

plot(data_shifted, label_time = TRUE) +
    geom_hline(yintercept = 0, linetype = "dotted")
```

Before shifting, the minimum (end of exercise) values for *smo2_left* and *smo2_right* were similar, but the starting baseline values were different. 

Whereas when we shift both baseline values to zero, we can see that the *smo2_right* signal has a smaller deoxygenation amplitude compared to the *smo2_left* signal.

We have to consider how our assumptions and processing decisions will influence our interpretations; by shifting both starting values, we are normalising for the starting condition of the two tissue sites and implicitly assuming that the baseline represents the same starting condition in both legs. 

For example, this may be appropriate when we are more interested in the relative change (delta) in each leg during an intervention or exposure (often referred to as `"∇SmO2"`).



## `rescale_mnirs()`

We may also want to rescale our data to a new dynamic range, changing the units to a new amplitude.

* `data`

    This function takes in a data frame, applies processing to all channels specified, then returns the processed data frame. *`mnirs`* metadata will be passed to and from this function.

* `nirs_channels`

    Channels should be grouped by providing a list (e.g. `list(c(A, B), c(C))`) where each group will be rescaled to a common range, and separate ranges between groups. The relative scaling between channels will be preserved within each group, but lost between groups.\
\
`nirs_channels` should be specified explicitly to ensure the intended grouping structure is returned. The default *`mnirs`* metadata will group all NIRS channels together.

* `range`

    Specifies the new dynamic range in the form `c(min, max)`. For example, if we want to calibrate each NIRS signal to their own observed 'functional range' during exercise, we could rescale them to 0-100%. 

```{r}
data_rescaled <- rescale_mnirs(
    data_filtered,    ## un-grouped nirs channels to rescale separately 
    nirs_channels = list(smo2_left, smo2_right), 
    range = c(0, 100) ## rescale to a 0-100% functional exercise range
)

plot(data_rescaled, label_time = TRUE) +
    geom_hline(yintercept = c(0, 100), linetype = "dotted")
```

Here, our assumption is that during a maximal exercise task, the minimum and maximum values represent the functional capacity of each tissue volume being observed. So we rescale the functional dynamic range in each leg.

By normalising this way, we might lose meaningful differences captured by the different amplitudes between *smo2_left* and *smo2_right*, but we might be more interested in the trend or time course of each response.

Our interpretation may be that *smo2_left* appears to start at a slightly higher percent of it's functional range, deoxygenates faster toward a minimum, and reaches a pseudo-plateau near maximal exercise. While *smo2_right* deoxygenates slightly slower and continues to deoxygenate until maximal task tolerance. Additionally, the right leg reoxygenates slightly faster than left during recovery. This might, for example, indicate exercise capacity differences between the limbs.



# 🔀 Pipe-friendly functions

Most *`mnirs`* functions can be piped together using Base R 4.1+ (`|>`) or `{magrittr}` (`%>%`). The entire pre-processing stage can easily be performed in a sequential pipe.

To demonstrate this, we'll read a different example file recorded with *Train.Red FYER* muscle oxygen sensor and pipe it through each processing stage straight to a plot.

This is also a good time to demonstrate how to use the global `mnirs.verbose` argument to silence all warning & information messages, for example when dealing with a familiar dataset. We recommend leaving `verbose = TRUE` by default whenever reading and exploring a new file.

```{r}
options(mnirs.verbose = FALSE)

read_mnirs(
    example_mnirs("train.red"),
    nirs_channels = c(
        smo2_left = "SmO2 unfiltered",
        smo2_right = "SmO2 unfiltered"
    ),
    time_channel = c(time = "Timestamp (seconds passed)"),
    zero_time = TRUE
) |>
    resample_mnirs() |> ## default will resample to fix irregular samples
    replace_mnirs(
        invalid_above = 73,
        outlier_cutoff = 3,
        span = 7
    ) |>
    filter_mnirs(
        method = "butterworth",
        order = 2,
        W = 0.01,
        na.rm = TRUE
    ) |>
    shift_mnirs(
        nirs_channels = list(smo2_left, smo2_right),
        to = 0,
        span = 60,
        position = "first"
    ) |>
    rescale_mnirs(
        nirs_channels = list(c(smo2_left, smo2_right)), ## 👈 channels grouped together
        range = c(0, 100)
    ) |>
    plot(label_time = TRUE)
```


# ✅ Signal Processing

After the NIRS signal has been cleaned and filtered, it should be ready for further processing and analysis. 

*`mnirs`* is being developed to include functionality for processing discrete intervals and events, e.g. reoxygenation kinetics, slope calculations for post-occlusion microvascular responsiveness, and critical oxygenation breakpoints.
