---
title: "interval-descriptive-params"
author:
    - name: Jem Arnold
date: last-modified
format:
    html:
        toc: true
        embed-resources: true
        output-dir: "."
        output-file: "interval-descriptive-params.html"
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
        warning: false
        message: false
        dpi: 300
        out.width: "100%"
        fig.align: "center"
---

## Aims

1. Read mNIRS file
2. Clean & pre-process
3. Extract intervals (baseline, deoxy, plateau, reoxy, hyperaemia)
4. Summarise descriptive parameters for each interval:
    a. baseline abs value
    b. deoxygenation rate
    c. deoxy half-time
    d. quasi-plateau slope
    e. quasi-plateau mean abs value
    f. minimum abs value
    g. reoxygenation rate
    h. reoxy half-time
    i. maximum abs value
    j. maximum abs amplitude (max - min)
5. Display group summary values

Examples:

![van Beekvelt et al., 2025 FIGURE 2. Schematic representation of the NIRS signals during the vascular occlusion test as well as the variables that were derived from this test. 1) Baseline saturation levels (BL) were calculated as the mean saturation level in percentage (%) over the last 30-s period prior to the start of the vascular occlusion. 2) Oxygen consumption (mVO 2 ) during rest was calculated from the rate of decrease in SmO 2 during the initial phase the occlusion, omitting the first 2 s to omit possible inflation artefacts. Linear regression was done over the subsequent 45-s period and expressed in %/s Maximum desaturation levels (VOT-min) during vascular occlusion were calculated as the mean desaturation over the last 30-s period of the 10-min occlusion and expressed as the relative change from baseline saturation levels. 4) Reperfusion rate (RR) following cuff release was calculated from the rate of increase in SmO 2 during the initial period of recovery, omitting the first 0.5 s to omit possible deflation artefacts. Linear regression was done over the subsequent 10-s period and expressed in %/s Maximum resaturation (HP-max) was calculated as the highest peak after cessation of the cuff and relative to the start of recovery after smoothing the recovery signal using a simple moving average with a window size of 100 samples, corresponding to approximately 10. DOI: 10.3389/fphys.2025.1659756](figures/van-Beekvelt-2025.jpg)

----

![Cornelis et al., 2021 Figure 1. Typical pattern of near infrared spectroscopy (NIRS) derived data during a treadmill walking test (unpublished data). Outcomes are categorised following the phase in the test: at rest, during exercise or deoxygenation, and during recovery or re-oxygenation. TSI ¼ tissue saturation index; MWD ¼ maximum walking distance. DOI: 10.1016/j.ejvs.2021.02.008](figures/Cornelis-2021.jpg)

----

```{r}
#| label: setup
#| code-fold: true
#| code-summary: "Show code"

library(tidyverse) ## for data wrangling
library(mnirs) ## for NIRS processing

options(mnirs.verbose = FALSE) ## silence warnings & info messages
theme_set(theme_mnirs()) ## set pretty default ggplot2 theme

```

```{r}
df <- read_mnirs(
    example_mnirs("train.red"),
    nirs_channels = c(smo2 = "SmO2 unfiltered"),
    time_channel = c(time = "Timestamp (seconds passed)"),
    zero_time = TRUE
) |>
    resample_mnirs() |> 
    filter_mnirs(method = "butter", order = 4, W = 0.01) |>
    extract_intervals(
        start = by_time(63),
        end = by_time(371),
        span = c(-60, 120)
    ) |>
    _[[1L]]

plot(df, label_time = TRUE)
```
