---
title: "Descriptive Interval Parameters"
author:
    - name: Jem Arnold
date: last-modified
format:
    html:
        toc: true
        embed-resources: true
        output-dir: "."
        output-file: "interval-params.html"
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
        warning: false
        message: false
        dpi: 300
        out.width: "100%"
        fig.align: "center"
---

## Aims

1. Read mNIRS file
2. Clean & pre-process
3. Extract intervals (baseline, deoxy, plateau, reoxy, hyperaemia)
4. Summarise descriptive parameters for each interval:
    - absolute values
        a. ✅ baseline mean
        b. ✅ quasi-plateau mean
        c. ✅ kquasi-plateau mean minimum
        d. ✅ hyperaemia mean maximum
        e. max exercise amplitude (max - min)
    - kinetics
        f. ✅ peak deoxygenation rate
        g. ❌ deoxy half-time
        h. ✅ mean quasi-plateau slope
        i. ✅ peak reoxygenation rate
        j. ❌ reoxy half-time
            - (❌ I don't quite have functions ready for this yet)
5. Display group summary values

## Literature example interval parameters

![van Beekvelt et al., 2025 FIGURE 2. Schematic representation of the NIRS signals during the vascular occlusion test as well as the variables that were derived from this test. 1) Baseline saturation levels (BL) were calculated as the mean saturation level in percentage (%) over the last 30-s period prior to the start of the vascular occlusion. 2) Oxygen consumption (mVO 2 ) during rest was calculated from the rate of decrease in SmO 2 during the initial phase the occlusion, omitting the first 2 s to omit possible inflation artefacts. Linear regression was done over the subsequent 45-s period and expressed in %/s Maximum desaturation levels (VOT-min) during vascular occlusion were calculated as the mean desaturation over the last 30-s period of the 10-min occlusion and expressed as the relative change from baseline saturation levels. 4) Reperfusion rate (RR) following cuff release was calculated from the rate of increase in SmO 2 during the initial period of recovery, omitting the first 0.5 s to omit possible deflation artefacts. Linear regression was done over the subsequent 10-s period and expressed in %/s Maximum resaturation (HP-max) was calculated as the highest peak after cessation of the cuff and relative to the start of recovery after smoothing the recovery signal using a simple moving average with a window size of 100 samples, corresponding to approximately 10. DOI: 10.3389/fphys.2025.1659756](figures/van-Beekvelt-2025.jpg)

----

![Cornelis et al., 2021 Figure 1. Typical pattern of near infrared spectroscopy (NIRS) derived data during a treadmill walking test (unpublished data). Outcomes are categorised following the phase in the test: at rest, during exercise or deoxygenation, and during recovery or re-oxygenation. TSI ¼ tissue saturation index; MWD ¼ maximum walking distance. DOI: 10.1016/j.ejvs.2021.02.008](figures/Cornelis-2021.jpg)

----

## Setup

```{r}
#| label: setup

library(tidyverse) ## for data wrangling
library(patchwork) ## for multi-plotting
library(mnirs)     ## for NIRS processing

theme_set(theme_mnirs()) ## set pretty default ggplot2 theme

```




## Read file, process, and extract mnirs intervals


In this example we will read one file with two work intervals.

We can use a pipe (base R >4.1 `|>`; or {magrittr} `%>%`) to chain reading and pre-processing operations with our function to extract the relevant intervals by lap. Further data cleaning and filtering can also be performed in this step, if needed.

Specifying two laps will extract a list of two data frames, for each interval. We can then feed this `data_list` into `purrr::map()`, which iterates operations over each item in a list. 

We want to add a `segment` column to the data, specifying where to demarcate our *"baseline"*, *"deoxy"*, *"plateau"*, and *"reoxy"* segments. 

We will have to manually find appropriate time cutoffs for capturing the deoxy and reoxy behaviour. During cycling exercise (in this example) I find 90-sec is a good default. This may get more complicated if deoxy/reoxy times are highly variable between trials.


```{r}
#| label: read & process mnirs intervals

## pre-define kinetics window timespan; adjust as needed
kinetics_span <- 90

## read example file with two intervals
df_list <- read_mnirs(
    file_path = example_mnirs("train.red"), ## example file path
    nirs_channels = c(smo2 = "SmO2"), ## detect and rename columns
    time_channel = c(time = "Timestamp (seconds passed)"),
    event_channel = c(lap = "Lap/Event"), ## laps identified as integers
) |>
    resample_mnirs() |> ## resample to regularise `time` values
    extract_intervals(
        start = by_lap(2, 4), ## extract intervals by lap number
        span = c(-60, 120), ## extend the interval bounds c(before, after)
        zero_time = TRUE ## set interval start time to `0`
    ) |>
    ## `purrr::map` iterates over a list;
    ## it operates over each list item (interval data frame)
    purrr::map(\(.df) {
        ## `time` values at interval start/end are contained in metadata
        ## view metadata via `attributes(data)`
        int_times <- attr(.df, "interval_times")

        ## create a new column with character factors for each segment
        ## some time cutoffs must be manually specified
        mutate(
            .df,
            segment = case_when(
                time <= int_times[1] ~ "baseline",
                between(
                    time,
                    int_times[1],
                    int_times[1] + kinetics_span
                ) ~ "deoxy",
                between(
                    time,
                    int_times[1] + kinetics_span,
                    int_times[2],
                ) ~ "plateau",
                time >= int_times[2] ~ "reoxy",
            )
        )
    })

## view the list of two interval data frames
df_list
```



## Plot intervals

The next step is to visualise the data to inspect for any issues. We will use `purrr::map()` again since we are still working with a list of two data frames.

We can easily plot side by side using `{patchwork}`.


```{r}
#| label: plot intervalsi

## iterate over our list of data frames
plot_list <- purrr::map(df_list, \(.df) {
    ## extract the interval times (start, end)
    ## add segments to capture the deoxy and reoxy segments
    int_times <- c(
        attr(.df, "interval_times"),
        attr(.df, "interval_times") + kinetics_span
    )

    ## built-in generic `mnirs.plot` with added vertical segment lines
    plot(.df, time_labels = TRUE) +
        geom_vline(xintercept = int_times, linetype = "dotted")
})

## `patchwork` allows joining plots with `+` to plot side by side
## or `/` to plot one on top of another
plot_list[[1]] + plot_list[[2]]
```

We can see the data look reasonable and the four segment demarcations are accurate.


## Summarise inteval parameters

Now we can calculate our summary parameters listed from the top of this document.

Currently, this is the janky part of the process. I don't yet have a wrapper function to return all of these values in a clean format. We will demonstrate how to use a few already existing {mnirs} internal functions. Internal package functions can be called using `mnirs:::`.

For local min-mean and max-mean SmO<sub>2</sub> values and peak slopes, we will use a 10-sec timespan as a local window. This can be modified as appropriate. 

Baseline and plateau segment means will include all data in the respective segment. A similar local rolling mean approach can be used if a specific timespan (e.g. 60-sec) is preferred for these values.

```{r}
#| label: summarise parameters

## iterate over our list of data frames, join summary data at the end
df_summary <- purrr::map(seq_along(df_list), \(.idx) {
    .df <- df_list[[.idx]]
    .int <- str_extract(names(df_list[.idx]), "\\d+") ## interval number

    ## find global mean-max and mean-min 10-sec SmO2 values
    ## using `mnirs` rolling functions to find local 10-sec means
    idx <- mnirs:::compute_local_windows(.df$time, span = 10)
    roll_mean <- mnirs:::compute_local_fun(
        .df$smo2, idx, mean, na.rm = TRUE
    )
    global_min <- min(roll_mean, na.rm = TRUE)
    global_max <- max(roll_mean, na.rm = TRUE)

    ## summarise per-segment mean SmO2, mean slopes, and peak slopes
    .df |>
        summarise(
            .by = segment,
            mean = mean(smo2, na.rm = TRUE),
            mean_slope = mnirs:::slope(smo2, time),
            peak_slope = mnirs:::peak_slope(smo2, time, span = 10)$slope,
        ) |>
        ## keep only the relevant metrics per segment, omit the rest
        mutate(
            mean = if_else(
                segment %in% c("baseline", "plateau"),
                mean,
                NA_real_
            ),
            mean_slope = if_else(
                segment == "plateau",
                mean_slope,
                NA_real_
            ),
            peak_slope = if_else(
                segment %in% c("deoxy", "reoxy"),
                peak_slope,
                NA_real_
            ),
        ) |>
        ## pivot from wide to long format to gather metrics by segment
        ## drop NA metrics
        pivot_longer(
            cols = c(mean, mean_slope, peak_slope),
            names_to = "metric",
            values_to = "value",
            values_drop_na = TRUE
        ) |>
        ## combine segment + metric names, stripping redundant slope qualifier
        ## e.g. deoxy_peak_slope -> deoxy_slope, plateau_mean_slope -> plateau_slope
        mutate(
            metric = str_remove(metric, "^(peak|mean)_(?=slope)"),
            name = paste(segment, metric, sep = "_")
        ) |>
        select(name, value) |>
        ## pivot from long back to wide 1-row df with new col names
        pivot_wider(names_from = name, values_from = value) |>
        mutate(
            min = global_min, ## attach global min/max scalars
            max = global_max,
            interval = .int, ## add interval id col
        ) |>
        relocate(interval, baseline_mean, min, max, plateau_mean)
}) |>
    list_rbind() ## bind list of two summary data frames into a single df

df_summary |>
    mutate(
        ## format numeric values for pretty printing
        across(matches(c("mean", "min", "max")), \(.x) {
            mnirs:::signif_trailing(.x, 1, "digits")
        }),
        across(matches("slope"), \(.x) {
            mnirs:::signif_trailing(.x, 3, "signif")
        }),
    ) |>
    ## print to document
    knitr::kable()    
```

This data frame could be written to file with `writexl::write_xlsx()` or `saveRDS()` for reproducible further processing.

The next step would be to iterate this entire process over a list of unique files, and combine the data frames into a longer list with individual participant id and load.


## Rescaling NIRS dynamic range

The same procedure can easily be done with NIRS data rescaled to the session dynamic range. I will show the process starting from scratch on the same example data.

The only different operation is calling `mnirs::rescale_mnirs()`, which scales the entire data range to the new specified min and max values. To instead scale the session to the occlusion ischaemic range, we would have to do something a bit more janky.


```{r}
#| label: normalised nirs signals

df_list <- read_mnirs(
    file_path = example_mnirs("train.red"),
    nirs_channels = c(smo2 = "SmO2"),
    time_channel = c(time = "Timestamp (seconds passed)"),
    event_channel = c(lap = "Lap/Event"),
) |>
    resample_mnirs() |>
    rescale_mnirs(range = c(0, 100)) |> ## rescale session min/max to 0/100
    extract_intervals(
        start = by_lap(2, 4),
        span = c(-60, 120),
        zero_time = TRUE
    ) |>
    purrr::map(\(.df) {
        int_times <- attr(.df, "interval_times")

        mutate(
            .df,
            segment = case_when(
                time <= int_times[1] ~ "baseline",
                between(
                    time,
                    int_times[1],
                    int_times[1] + kinetics_span
                ) ~ "deoxy",
                between(
                    time,
                    int_times[1] + kinetics_span,
                    int_times[2],
                ) ~ "plateau",
                time >= int_times[2] ~ "reoxy",
            )
        )
    })

## let's plot it to view the y-axis values
plot_list <- purrr::map(df_list, \(.df) {
    int_times <- c(
        attr(.df, "interval_times"),
        attr(.df, "interval_times") + kinetics_span
    )

    plot(.df, time_labels = TRUE) +
        scale_y_continuous(limits = c(0, 100), expand = expansion(0)) +
        geom_vline(xintercept = int_times, linetype = "dotted")
})

plot_list[[1]] + plot_list[[2]]
```

Note the expanded y-axes range, but the relative scaling is the same as before.



```{r}
#| label: turn it into a function
#| echo: false

extract_parameters <- function(df_list) {
    purrr::map(seq_along(df_list), \(.idx) {
        .df <- df_list[[.idx]]
        .int <- str_extract(names(df_list[.idx]), "\\d+") ## interval number

        ## find global mean-max and mean-min 10-sec SmO2 values
        ## using `mnirs` rolling functions to find local 10-sec means
        idx <- mnirs:::compute_local_windows(.df$time, span = 10)
        roll_mean <- mnirs:::compute_local_fun(
        .df$smo2, idx, mean, na.rm = TRUE
    )
        global_min <- min(roll_mean, na.rm = TRUE)
        global_max <- max(roll_mean, na.rm = TRUE)

        ## summarise per-segment mean SmO2, mean slopes, and peak slopes
        .df |>
            summarise(
                .by = segment,
                mean = mean(smo2, na.rm = TRUE),
                mean_slope = mnirs:::slope(smo2, time),
                peak_slope = mnirs:::peak_slope(smo2, time, span = 10)$slope,
            ) |>
            ## keep only the relevant metrics per segment, omit the rest
            mutate(
                mean = if_else(
                    segment %in% c("baseline", "plateau"),
                    mean,
                    NA_real_
                ),
                mean_slope = if_else(
                    segment == "plateau",
                    mean_slope,
                    NA_real_
                ),
                peak_slope = if_else(
                    segment %in% c("deoxy", "reoxy"),
                    peak_slope,
                    NA_real_
                ),
            ) |>
            ## pivot from wide to long format to gather metrics by segment
            ## drop NA metrics
            pivot_longer(
                cols = c(mean, mean_slope, peak_slope),
                names_to = "metric",
                values_to = "value",
                values_drop_na = TRUE
            ) |>
            ## combine segment + metric names, stripping redundant slope qualifier
            ## e.g. deoxy_peak_slope -> deoxy_slope, plateau_mean_slope -> plateau_slope
            mutate(
                metric = str_remove(metric, "^(peak|mean)_(?=slope)"),
                name = paste(segment, metric, sep = "_")
            ) |>
            select(name, value) |>
            ## pivot from long back to wide 1-row df with new col names
            pivot_wider(names_from = name, values_from = value) |>
            mutate(
                min = global_min, ## attach global min/max scalars
                max = global_max,
                interval = .int, ## add interval id col
            ) |>
            relocate(interval, baseline_mean, min, max, plateau_mean)
    }) |>
        list_rbind() ## bind list of two summary data frames into a single df
}
```


## Summarise rescaled interval parameters

I've secretly wrapped the janky `summarise()` operation we used above into a function, so we can call it quickly again here:


```{r}
#| label: summarise rescaled params

## repeat the same summarise operation and display the table
extract_parameters(df_list) |>
    mutate(
        ## format numeric values for pretty printing
        across(matches(c("mean", "min", "max")), \(.x) {
            mnirs:::signif_trailing(.x, 1, "digits")
        }),
        across(matches("slope"), \(.x) {
            mnirs:::signif_trailing(.x, 3, "signif")
        }),
    ) |>
    ## print to document
    knitr::kable()
```

Compare value and slope differences after rescaling. Carefully consider implications and assumptions of rescaling or not, or the data provided as agreement values. Consider what values will NIRS practitioners and users be viewing on a daily basis? Rescaled or unscaled?

Relative reliability will not be affected, but absolute agreement will.