---
title: "Analysing muscle oxidative capacity with {mnirs}"
author:
    - name: Jem Arnold
date: last-modified
format:
    html:
        embed-resources: true
        toc: true
bibliography: oxcap-citations.bib
csl: ../apa.csl
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
        dpi: 300
        out.width: "100%"
        fig.align: "center"

---

# Introduction

The [`{mnirs}`](https://github.com/jemarnold/mnirs) package is under active development with processing and analysis methods for muscle near-infrared spectroscopy in R.

In this article, I will demonstrate how some recently added functionality can be combined to perform muscle oxidative capacity (OxCap) analysis from a repeated ischaemic occlusion protocol, while ensuring reproducibility and adhering to current gold-standard processing methods.

::: {.callout-note}
This article assumes basic familiarity with the {mnirs} package. For an overview and demonstration of data processing with {mnirs}, please see the original package vignette [*Reading and Cleaning Data with {mnirs}*](https://jemarnold.github.io/mnirs/articles/reading-mnirs-data.html).
:::

::: {.callout-tip}
Skip to the [analysis code](#analysis-plan), below.
:::

## Muscle oxidative capacity testing

One of the most compelling applications emerging in mNIRS research is the ability to non-invasively evaluate gross **muscle oxidative capacity** after a dynamic exercise task.

Oxidative capacity is the maximal rate at which a muscle utilises oxygen (O<sub>2</sub>) to meet the energetic demand of exercise, and is related to mitochondrial respiratory function [@Beever2020].

Traditionally, OxCap assessment requires highly complex and invasive in-vitro methods such as high-resolution respirometry of biopsy tissue samples, or expensive in-vivo <sup>31</sup>P magnetic resonance spectroscopy.

With mNIRS, muscle OxCap can be assessed non-invasively during more dynamic exercise tasks, and with lower cost and burden than traditional methods previously available.

## Protocol overview

The participant is asked to perform a brief exercise task or muscle contraction stimulus to elevate the O<sub>2</sub> demand in the target tissue, with mNIRS sensors over the peripheral muscle of interest and an occlusion cuff ready (deflated) around the proximal limb.

Immediately after the stimulus, the occlusion cuff is rapidly inflated to a supra-systolic blood pressure (e.g., 300 mmHg) to transiently stop blood flow into the distal target muscle site.

The occlusion is held for a brief period such as 5-seconds, then deflated to allow recovery to proceed. A sequence of these brief, repeated occlusions are performed at a pre-specified tempo (e.g. 5-sec occlusion, 10-sec recovery) for up to twenty repetitions (~5-minutes) until the muscle has recovered back to baseline.

![Figure 2 from @McCully2024. (A) The experimental setup for the progressive exercise test. The NIRS device was placed on the vastus lateralis using straps. The blood pressure cuff was placed proximal to the NIRS device. The leg is shown lifting the weight. For NIRS measurements, the leg was placed on top of the padding in the horizontal position to allow the muscle to relax during the measurements.](figures/McCully-2024.jpg)

During each occlusion oxygen delivery is restricted and assumed to be zero. Therefore, the rate of deoxygenation — the slope of the rise in deoxy[haem] (HHb/time; μM/sec), or decline in oxygen saturation (SmO<sub>2</sub>) or oxy[haem] — is interpreted to represent the rate of local muscle oxygen uptake (mV̇O<sub>2</sub>).

Immediately after the exercise task when metabolic demand is high, mV̇O<sub>2</sub> will be elevated. It will decline at an exponential rate as recovery proceeds back to baseline. The rate constant (*k*, min<sup>-1</sup>) of this monoexponential curve quantifies the muscle OxCap value [@Adami2018].

![Figure 1 from @Adami2018. Muscle oxygen consumption (mVO2) recovery rate constant (k) by near-infrared spectroscopy. A and B: example of the oxidative capacity test by NIRS. A: changes in the tissue saturation index (TSI) during dynamic exercise (EX, gray area) and subsequent intermittent arterial occlusions at rest. B: mVO2 recovery kinetics derived from the rate of change of TSI during intermittent arterial occlusions measured from A. mVO2 recovery data are fit to an exponential (dashed line) to estimate the recovery k. C: summaries of current reports of the mVO2 recovery rate constant (k), which is proportional to oxidative capacity, in upper and lower limbs of adults in health and disease.](figures/Adami-2018.jpg)

This technique has been well studied in recent years, with validation against <sup>31</sup>P-MRS [@Ryan2013] and mitochondrial content protein markers [@Tripp2023]. The rate constant *k* has been shown to be proportionally higher (faster recovery) in endurance-trained muscle, and lower (slower) in untrained, older, or diseased muscle [@Adami2018].



## A call for standardised processing methods

Technical procedures and analysis methods for OxCap assessment have converged toward standardisation in recent years, with methods showing good reliability and reproducibility.

Recently there has been a call to implement standard analysis scripts with robust slope detection and non-linear modelling, to minimise operator-related variability between research centres, and thereby improving interpretation of outcomes across populations and interventions [@Costalat2025; @Rasica2024].

I wanted to see how well (or poorly) my current {mnirs} functionality could process and analyse repeated occlusion muscle OxCap data.

In the future, this process could be wrapped into a single convenience function to make the process easier for the end user. But I believe this process is already considerably simpler and more robust than non-reproducible methods relying on manual data cleaning, manual occlusion interval selection, and iterative Excel-macro curve fitting.



## Analysis plan

For OxCap analysis, we will perform roughly 8 steps:

1. Import an example mNIRS file with the repeated occlusions procedure.
2. Process/clean data, as needed (minimal in this example).
3. Iteratively correct NIRS values for changes in blood volume, a required step to ensure valid analysis.
4. Specify occlusion events in the data and extract a 5-sec interval for each occlusion.
5. Find the peak 3-sec linear regression NIRS slope within each occlusion interval.
6. Fit a monoexponential curve across NIRS slope observations for both repeated occlusion trials.
7. Extract the rate constant (*k*) for each trial exponential curve.
8. Plot the modelled data.



# Oxidative capacity analysis with {mnirs}

This article serves as a vignette for some of the kinetics analysis functions recently added to the {mnirs} package, and how they can be used together to perform robust OxCap analysis.

* `correct_blood_volume()` is used to normalise values for oxy[haem], deoxy[haem], and total[haem] channels (if present), using an iterative method that properly accommodates negative values, developed from @Beever2020. See `?correct_blood_volume` for details.

* `extract_intervals()` is used to detect specific events or times in an *"mnirs"* data frame, extract an interval around each one, and return a list of data frames.\
\
This is useful when a certain label is given in `event_channel`, or a list of time values can be provided manually for when events occur, such as the start of an exercise interval or occlusion. The returned list of data frames is ready for further analysis. See `?extract_intervals` for details.

* `analyse_kinetics()` is used to evaluate mNIRS response kinetics from a single *"mnirs"* data frame or a list of data frames, with a selection of parametric and non-parametric methods. It returns a formatted table of model results. Internal model components can be extracted with `attributes(results)`. See `?analyse_kinetics` for details.\
\
We will use this function twice for OxCap analysis:

  * `analyse_kinetics(method = "peak_slope")` is used to find the peak linear regression slopes on deoxy[haem] from the extracted occlusion intervals. The steepest (positive or negative) slope is often interpreted as the point of greatest mismatch between O<sub>2</sub> supply and O<sub>2</sub> extraction, such as for muscle OxCap testing. See `?peak_slope` for more details on the vector-level function called by `analyse_kinetics()`.

  * `analyse_kinetics(method = "monoexponential")` is then used to fit an exponential curve through the resulting slope values over time, to find the rate constant `k`. A self-starting function `SSmonoexponential()` is fit with `stats::nls()` and can be called as either a 4-parameter monoexponential model with a time-delay parameter (TD), or a reduced 3-parameter model without TD where the onset of the response is expected to be immediate. See `?monoexponential` and `?SSmonoexponential` for more details.



## Setup

First, load our packages and initial setup options.

```{r}
#| label: setup
#| warning: false

library(dplyr) ## for convenient data wrangling
library(ggplot2) ## for plotting
library(mnirs)

options(mnirs.verbose = FALSE) ## globally silence package messages
theme_set(theme_mnirs()) ## set ggplot2 theme
```


::: {.callout-note title="Acknowledgement"}
I received this example file from [Dr. Thomas Tripp](https://scholar.google.com/citations?user=A9JwQbYAAAAJ&hl=en), currently a postdoctoral fellow in [Dr. Martin MacInnis](https://scholar.google.com/citations?hl=en&user=tJepNIQAAAAJ)' lab at the University of Calgary. They have kindly allowed me to include this file with the {mnirs} package for users to examine themselves.

It can be accessed by calling `example_mnirs("portamon-oxcap")`.
:::



# Read {mnirs} data file

The included example file has three NIRS channels and a column with sample number, which will automatically be converted to a time value when the `read_mnirs()` function recognises the *Artinis Oxysoft* export format and sample rate (10 Hz). The file does not need to be cleaned for outliers or digitally filtered, so minimal pre-processing is required.

One thing to note: by default, *Artinis Oxysoft* exports an event label column without a named header. This is the column we will use to identify our occlusion events by the label *"Occlusion"*. In {mnirs}, an unnamed column like this will be coerced to a name `"col_*"` with a numeric suffix for the column number in the data file. In our example file, this events column will therefore be `"col_6"`. It's not the most elegant solution, but it works for now.


```{r}
#| label: read file and process

df <- read_mnirs(
    file_path = example_mnirs("portamon-oxcap.xlsx"),
    nirs_channels = c(thb = 2, hhb = 3, o2hb = 4), ## identify and rename channels
    event_channel = c(event = "col_6") ## specify the events column
)

## view the structure of our data frame
df
```

This file has two sequential occlusion trials. There are many different ways to identify & separate trials. We will find the long gap between consecutive occlusion labels and split the data frame at that point.

```{r}
#| label: split df

## find time gap between occlusion trials
occl_times <- df$time[grepl("Occlusion", df$event)]
split_time <- occl_times[which(diff(occl_times) > 60)] + 10

plot(df) +
    geom_vline(xintercept = split_time, linetype = "dashed")
```


The two exercise bouts and repeated occlusion trials can clearly be identified from plotting the raw data. A few resting occlusions were also performed at the start of the file, but we will ignore those and just model the two repeated occlusion trials.



# Correct for blood volume

A necessary processing step before analysing the OxCap trials is to apply a correction to the NIRS channels for changes in blood volume during the repeated occlusions. This step has been shown to improve validity when taking slopes of the oxy- or deoxy[haem] channels [@Ryan2012].

The preferred method is to iteratively correct for instantaneous changes in oxygenation between samples, according to @Beever2020. `correct_blood_volume()` uses this method, modified to accommodate negative NIRS values, which can be reported by some mNIRS devices.

## `correct_blood_volume()`

* `data`

    This function takes in a data frame, corrects the specified NIRS channels for blood volume, and returns the data frame with the corrected `nirs_channels`.\
\
At least two of the three channels must be named explicitly as `oxy_channel`, `deoxy_channel`, and `total_channel`; the third is derived. These cannot be read from *"mnirs"* metadata, to avoid ambiguity. The corrected channels are passed on in metadata as `nirs_channels`.


```{r}
#| label: blood volume correction

df <- df |>
    correct_blood_volume(
        oxy_channel = o2hb,
        deoxy_channel = hhb,
        total_channel = thb
    )

## plot corrected data with occlusion event indicators
plot(df) +
    geom_vline(xintercept = occl_times, linetype = "dotted", alpha = 0.4)
```


This results in a symmetrical distribution of oxy- and deoxy[haem], with the sum total (total[haem]; THb) definitionally equal to zero at all samples.

We can also plot occlusion event indicators (vertical dotted lines) to visually confirm timing and detect any erroneous events.


# Extract occlusion intervals

The next step is to identify these occlusion events and extract each interval into a list of data frames which we can continue to analyse iteratively. We can run this function twice to easily separate the occlusions within the two trial.


## `extract_intervals()`

* `data`

    This function takes in a data frame or list of data frames, and returns a list of data frames, named per detected interval. *"mnirs"* metadata is passed to and from this function.

* `nirs_channels`, `time_channel`, `event_channel`, & `sample_rate`

    Specify which column names in `data` will be processed. `nirs_channels` are the response variables; `time_channel` is the time series predictor variable; `event_channel` specifies where to look for event labels; and `sample_rate` carries the measurement/export samples-per-second rate used when ensemble-averaging across intervals with unequal time samples. If any arguments are not specified, they will be retrieved from *"mnirs"* metadata. Channels in the data not explicitly specified will be passed through unprocessed to the returned data frames.\
\
We will analyse OxCap from the deoxy[haem] (HHb) NIRS signal, as often recommended [@Costalat2025].
    
* `start` & `end`

    Interval boundaries are specified using helper functions: `by_time()` for `time_channel` values; `by_label()` for `event_channel` labels; `by_lap()` for lap indices in `event_channel`; or `by_sample()` for sample indices (row numbers). Provide both `start` and `end` to define precise intervals, or provide `start` alone with `span` to extract windows around events.\
\
First, we can use `start` and `end` to split the data frame between two trials. Then, since the occlusions are all identified by the event label *"Occlusion"*, it is very simple to detect this string in the `event` column as the `start` value to extract each interval.\
\
Detecting intervals by event labels must match a full string exactly or use a regular expression (regex) string such as *"(?i)occlusions?"* or *"start|begin"*.

* `group_intervals`

    Multiple events can be extracted for analysis as `"distinct"` intervals, or `"ensemble"`-averaged together. Different groupings of intervals can be ensemble-averaged by providing a list of intervals by number; e.g. `list(c(1, 2), c(3, 4))`.\
\
For manual grouping, the interval numbers need to be known ahead of time and always refer to their order of occurrence in the data frame (i.e., sorted by `time_channel`). Any event intervals detected but not supplied by number will be included as distinct data frames.\
\
For this example, we want to analyse each interval separately using `group_intervals = "distinct"`.

* `span`

    A two-element vector of `time_channel` values, specifying a time window as `c(before, after)`, where positive values indicate time after the event sample and negative values indicate time before the event. When both `start` and `end` are provided, `span` shifts boundaries additively: `span[1]` adjusts starts, `span[2]` adjusts ends.\
\
ince all occlusion bouts were held for 5-seconds in this protocol, we can use a 5-sec time span. If the occlusion duration changed through the protocol, we could specify a list of time spans for each (`span = list()`). In this case we can provide a single vector which will be recycled.\
\
From visual investigation of the plots (not shown), we noticed that the deoxy[haem] signal was disrupted by the occlusion cuff inflation during the first ~0.5 sec of each occlusion interval. This might influence slope values, so we will exclude the first second by specifying `span = c(1, 5)`, i.e. include from second 1 (rather than 0) to 5 after each *"Occlusion"* event label. To adjust the start boundary earlier, we could specify a negative `span` value, e.g. `span = c(-2.5, 2.5)` to include a 5-sec span centred on the label itself.

* `zero_time`

    The time values for each event interval can be recalculated to start from zero at the event indicator. Time values for ensemble-averaged intervals will always be re-calculated from zero, since the original time information is lost when ensembling.\
\
For repeated occlusion testing, the original time information for each occlusion interval is required to subsequently model the exponential recovery function, so we keep `zero_time = FALSE`.


```{r}
#| label: extract intervals
#| fig-width: 9
#| fig-asp: 0.8

## split the df at the identified time value
extract_intervals(
    df,
    group_intervals = "distinct", ## return the two trials separately
    start = by_time(c(0, split_time)), ## start times at the start and split
    end = by_time(split_time, tail(df$time, 1)), ## end times at the split and end
    span = c(0, 0) ## no additional buffer around the start/end boundaries
) |>
    ## pipe the list of two trial dfs into another `extract_intervals()`
    ## identify and extract intervals by event label "Occlusion"
    extract_intervals(
        group_intervals = "distinct", ## return each occlusion event separately
        start = by_label("occlusion", ignore_case = TRUE),
        span = c(1, 5),   ## interval c(start, end) around each event time value
        zero_time = FALSE ## preserve original time values at each event
    ) -> df_list          ## right-assign list of all occlusion intervals

## visualise facet plot with all occlusion intervals for error checking
## the steepest slope facets are the first occlusions from each trial
plot(df_list, ncol = 8) +
    theme(axis.text.x = element_blank())
```


We can plot the sequence of extracted occlusion intervals to visually inspect the data. As expected, the first interval during each repeated occlusion trial appears the steepest, with the slope rapidly becoming less steep during each subsequent occlusion as mV̇O<sub>2</sub> recovers.

From here, we want to calculate the slope values for each of these occlusion intervals.


# Calculate occlusion slopes


## `analyse_kinetics()`

* `data`

    This function takes in a data frame or a list of data frames, analyses all channels within each data frame separately, and returns a formatted table of results. Model components such as coefficients, fitted data frames, input metadata for each channel, and model diagnostics can be retrieved from results metadata with `attributes()` or `names()`.

* `nirs_channels` & `time_channel`

    If the data contain *"mnirs"* metadata, these will be detected automatically, or they can be specified explicitly for each data frame, or a single vector of channel names will be recycled for all data frames.

* `start_time`, `end_window` & `direction`

    These arguments apply to every method. `start_time` marks the response onset, where the fit time is effectively zero. When intervals come from `extract_intervals()`, it is retrieved automatically from *"mnirs"* metadata; otherwise it defaults to the first time value. In this example, each interval already has a start time of zero, so we can leave this blank.\
\
`end_window` defines the end of the fitting window by looking forward for a reasonable end time for the kinetics response, by taking the first peak or trough with no more extreme value beyond that time span. The default `end_window = Inf` uses the global extreme across the whole interval, which is appropriate for these short occlusion windows.\
\
`direction` (default `"auto"`) detects whether the response rises or falls; it can be set explicitly to `"positive"` or `"negative"`. For parametric fits it also constrains the sign of the fitted amplitude `B − A`.\
\
For processing deoxy[haem], we specify `direction = "positive"`, since deoxygenation increases during occlusions.

* `method`

    Four analysis methods are currently available. For this OxCap analysis, we will need to use two of them: *"peak_slope"* and *"monoexponential"*.



### Analyse peak slope kinetics

* `method = "peak_slope"`

The **peak slope** method is a *semi-parametric* approach: it fits a linear regression model across a sliding local window, returning the single steepest slope along with the time at which it occurs. For OxCap analysis, the steepest local rate of deoxygenation is interpreted as the point of greatest mismatch between O<sub>2</sub> delivery and extraction — a proxy for mV̇O<sub>2</sub> during each occlusion.

* `width`, `span`

    The local window is specified by one of either `width`, a number of samples, or `span`, a range of time in units of `time_channel`.\
\
For this occlusion protocol, we care about the peak 3-sec slope value from each occlusion interval.

* `align`

    The window can be `"centre"`-aligned around the target sample (`idx`), or `"left"`-aligned with `idx` at the start of the window (forward-looking), or `"right"`-aligned with `idx` at the end (backward-looking).\
\
We will keep the default `"centre"` alignment.

* `partial`

    By default, a local window will only return a linear regression model if all samples within the window are valid. If `partial` is set to `TRUE`, then local windows will return a slope model as long as there are two or more samples for a valid fit.\
\
A window with fewer samples will tend to return steeper slope values, which may or may not be relevant to our interpretations. e.g., as an edge-case (literally at the edges of the data) in a noisy data set there may be two samples with an extreme local slope between them. We usually would want to ignore this outlier slope value.


```{r}
#| label: find peak slopes
#| warning: false
#| fig-width: 9
#| fig-asp: 0.8

## analyse local peak slopes within each occlusion interval
slope_results <- analyse_kinetics(
    df_list,
    nirs_channels = hhb,
    time_channel = time,
    method = "peak_slope",
    span = 3,
    direction = "positive",
    partial = FALSE,
)

## view the formatted results
slope_results

plot(slope_results, ncol = 8, label_size = 3.5) +
    theme(axis.text.x = element_blank())
```

We can quickly visualise the individual kinetics results with the built-in `plot()` function (see `?plot.mnirs_kinetics` for details). But we will build a custom plot to better visualise the exponential response kinetics for mV̇O<sub>2</sub> in each trial.

Next, we will extract the slope values into their own data frame to plug into the kinetics analysis.


```{r}
#| label: create slopes_df

## extract vector of interval start times from kinetics metadata
interval_times <- slope_results$interval_times$start_times

## extract a data frame with slope coef & time values, grouped by trial
slopes_df <- slope_results$coefficients |>
    mutate(
        trial = factor(sub("^[^_]+_(\\d+).*", "\\1", interval)), ## add trial by regex
        ## return centre of linear slope within each occlusion interval
        time = peak_slope_time + interval_times,
        peak_slope_time = NULL,
    ) |>
    group_by(trial) |>
    ## recalculate time starting from zero for each trial
    mutate(time = time - time[1L]) |>
    select(trial, time, slope)
```


```{r}
#| label: plot peak slopes

## plot observed slope values for each trial
p <- ggplot(slopes_df, aes(x = time, y = slope, colour = trial)) +
    coord_cartesian(ylim = c(0, NA)) +
    scale_x_continuous(
        name = "Time (mm:ss)",
        breaks = breaks_timespan(),
        labels = format_hmmss
    ) +
    scale_y_continuous(
        name = expression(bold(HHb ~ Slope ~ '(μM' %.% sec^'-1' * ')')),
    ) +
    scale_colour_mnirs() +
    geom_line() +
    geom_point(fill = "white", size = 3, shape = 21, stroke = 1)

p
```



# Model mV̇O<sub>2</sub> recovery


## `analyse_kinetics()`


### Analyse monoexponential kinetics

* `method = "monoexponential"`

The **monoexponential** method is a *parametric* approach: it fits a self-starting exponential curve to time series data with `nls()`, using `SSmonoexponential()` to initialise. This models how mV̇O<sub>2</sub> recovers back to baseline after the exercise stimulus, and its rate constant *k* is the OxCap outcome.

It fits either a **4-parameter** model `A + (B − A)(1 − exp(−(t − TD) / tau))` with a time delay *TD*, or a reduced **3-parameter** model without *TD* when the response is expected to start immediately.

* `use_TD` (*TD* coefficient)

    This is the only argument specific to *"monoexponential"* method, and explicitly specifies a 4-parameter monoexponential function with a time delay between the stimulus and the start of a systematic exponential response. This captures behaviour which does not respond instantaneously to an intervention with a delay or a transient acceleration phase before exponential behaviour. The model will fall back to attempt a 3-parameter fit if the 4-parameter fails to converge.\
\
Often during repeated occlusions after exercise, the first occlusions may not conform to an exponential response if the oxygen saturation is too low and limiting to mV̇O<sub>2</sub>. In these cases, a time delay parameter may be used to accurately capture the true monoexponential response. This protocol was designed to avoid this potential delay and so a time delay parameter was not required.

* *A*, *B* coefficients

    The starting (baseline) value and the ending (asymptote) value, respectively for the monoexponential curve. Will define either an exponential association curve in the positive direction when *B* > *A*, or exponential decay in the negative direction when *A* > *B*. 

* *tau* coefficient

    The time constant (τ) describes the speed of the exponential response from *A* to *B*, and is approximated by the time elapsed at 63.2% of the total response amplitude. i.e., when the predicted value `y` is equal to `A + 0.632 × (B − A)`.\
\
The rate constant (*k*) is the inverse of the time constant (`k = 1 / tau`) and is also commonly used in NIRS research, including for OxCap assessment. *k* is returned as the inverse of `time_channel` units (i.e., seconds) by default. In this example we will convert *k* to min<sup>-1</sup>.


```{r}
#| label: monoexponential curve modelling

## use 3-param monoexponential model for each trial
## fit models within the data frame grouped by trial
monoexp_results <- analyse_kinetics(
    slopes_df,
    nirs_channels = slope,
    time_channel = time,
    method = "monoexponential",
    use_TD = FALSE ## expect immediate response onset
)

## view the formatted results
monoexp_results
```

```{r}
#| label: build coef_labels

## extract and bind the list of fitted data for plotting
fitted_data <- do.call(rbind, monoexp_results$data)

## extract tau and k coef labels per trial, for plotting
coef_labels <- monoexp_results$coefficients |>
    rename(trial = interval) |>
    summarise(
        .by = trial,
        label = sprintf(
            "Trial %s\ntau = %s sec\nk = %s min⁻¹",
            trial,
            round(tau, 1),
            round(k * 60, 1) ## convert to min^-1
        )
    )
```

```{r}
#| label: plot modelled data

## add predicted values to the plot of observed data
p +
    geom_line(
        data = fitted_data,
        aes(y = slope_fitted, group = trial),
        linewidth = 0.8, colour = "black"
    ) +
    geom_text(
        data = coef_labels, show.legend = FALSE,
        aes(label = label, colour = trial, group = trial),
        x = Inf, y = Inf,
        size = 5, hjust = 1.1, vjust = c(1.5, 3)
    )
```

# Conclusion

This article demonstrates how muscle oxidative capacity analysis can be performed using {mnirs} with some relatively simple, reproducible data wrangling steps.

In future development, I would like to wrap these analysis steps into a dedicated convenience function with recommended default parameters, to make the analysis process more streamlined.

However, there are always trade-offs between automation and customisation: fitting nonlinear models with relatively sparse data samples (there are only samples per trial in this example) with real-world noisy biological data can result in unstable fit parameters or inability to converge at all. Manual processing steps may be inevitably required during analysis to accommodate for these automation limitations.

# References