---
title: "Analysing muscle oxidative capacity with {mnirs}"
author:
    - name: Jem Arnold
date: last-modified
format:
    html:
        embed-resources: true
        toc: true
bibliography: oxcap-citations.bib
csl: ../apa.csl
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
        dpi: 300
        out.width: "100%"
        fig.align: "center"

---

# Introduction

The [`{mnirs}`](https://github.com/jemarnold/mnirs) package is under active development with processing and analysis methods for muscle near-infrared spectroscopy in R.

This article demonstrates recently added functionality in {mnirs} can be combined to perform muscle oxidative capacity (OxCap) analysis from a repeated ischaemic occlusion protocol, ensuring reproducibility and adhering to current gold-standard processing methods.

::: {.callout-note}
This article assumes basic familiarity with the {mnirs} package. For an overview and demonstration of data processing with {mnirs}, please see the original package vignette [*Reading and Cleaning Data with {mnirs}*](https://jemarnold.github.io/mnirs/articles/reading-mnirs-data.html).
:::

::: {.callout-tip}
Skip to the [analysis code](#oxidative-capacity-analysis-with-mnirs), below.
:::

## Muscle oxidative capacity testing

One of the most compelling applications emerging in mNIRS research is the non-invasive evaluation of gross **muscle oxidative capacity** after a dynamic exercise task.

Oxidative capacity is the maximal rate at which a muscle utilises oxygen (O<sub>2</sub>) to meet the energetic demand of exercise, and is related to mitochondrial respiratory function [@Beever2020].

Traditional mitochondrial oxidative capacity assessment requires invasive in-vitro methods such as high-resolution respirometry of biopsy tissue, or expensive in-vivo <sup>31</sup>P magnetic resonance spectroscopy. mNIRS is comparatively far more accessible, and allows non-invasive assessment during dynamic exercise tasks, at lower cost and with lower participant burden.

## Protocol overview

The participant performs a brief exercise task or muscle stimulation to elevate O<sub>2</sub> extraction (mV̇O<sub>2</sub>) demand in the target tissue, with mNIRS sensors over the muscle of interest and an occlusion cuff around the proximal limb.

Immediately after the stimulus, the cuff is rapidly inflated to a supra-systolic pressure (e.g. 300 mmHg), transiently stopping blood flow into the distal target muscle. The occlusion is held briefly (e.g. 5 s) then rapidly deflated to allow recovery. A sequence of these brief, repeated occlusions are performed at a pre-specified tempo (e.g. 5-sec occlusion, 10-sec recovery) for up to twenty repetitions (~5 min), until the muscle recovers to baseline.

![Figure 2 from @McCully2024. (A) The experimental setup for the progressive exercise test. The NIRS device was placed on the vastus lateralis using straps. The blood pressure cuff was placed proximal to the NIRS device. The leg is shown lifting the weight. For NIRS measurements, the leg was placed on top of the padding in the horizontal position to allow the muscle to relax during the measurements.](figures/McCully-2024.jpg)

During each occlusion, oxygen delivery is restricted and assumed to be zero. The rate of deoxygenation — the slope of the rise in deoxy[haem] (HHb/time; μM/sec), or decline in oxygen saturation (SmO<sub>2</sub>) or oxy[haem] — is therefore interpreted as the rate of local muscle oxygen uptake (mV̇O<sub>2</sub>).

mV̇O<sub>2</sub> is elevated immediately after the exercise task, then recovers exponentially toward baseline. The rate constant (*k*, min<sup>-1</sup>) of this monoexponential curve quantifies the muscle OxCap value [@Adami2018].

![Figure 1 from @Adami2018. Muscle oxygen consumption (mVO2) recovery rate constant (k) by near-infrared spectroscopy. A and B: example of the oxidative capacity test by NIRS. A: changes in the tissue saturation index (TSI) during dynamic exercise (EX, gray area) and subsequent intermittent arterial occlusions at rest. B: mVO2 recovery kinetics derived from the rate of change of TSI during intermittent arterial occlusions measured from A. mVO2 recovery data are fit to an exponential (dashed line) to estimate the recovery k. C: summaries of current reports of the mVO2 recovery rate constant (k), which is proportional to oxidative capacity, in upper and lower limbs of adults in health and disease.](figures/Adami-2018.jpg)

This technique is well validated against <sup>31</sup>P-MRS [@Ryan2013] and mitochondrial content protein markers [@Tripp2023]. The rate constant *k* is proportionally higher (faster recovery) in endurance-trained muscle, and lower (slower) in untrained, older, or diseased muscle [@Adami2018].



## A call for standardised processing methods

Technical and analysis methods for OxCap assessment have converged toward standardisation in recent years, with good demonstrated reliability and reproducibility. There has been a recent call to implement standard analysis scripts with robust slope detection and non-linear modelling, to minimise operator-related variability between research centres, and improve interpretation of outcomes across populations and interventions [@Costalat2025; @Rasica2024].

This article uses recently developed {mnirs} functionality to process and analyse repeated occlusion muscle OxCap data. In the future, this could be wrapped into a single convenience function, but the current process is already considerably simpler and more robust than methods relying on manual data cleaning, manual interval selection, and iterative (e.g. Excel macro) curve fitting.



## Analysis plan

For OxCap analysis, we will perform roughly 8 steps:

1. Import an example mNIRS file with the repeated occlusions procedure.
2. Process/clean data, as needed (minimal in this example).
3. Iteratively correct NIRS values for changes in blood volume, a required step to ensure valid analysis.
4. Specify occlusion events in the data and extract a 5-sec interval for each occlusion.
5. Find the peak 3-sec linear regression NIRS slope within each occlusion interval.
6. Fit a monoexponential curve across NIRS slope observations for both repeated occlusion trials.
7. Extract the rate constant (*k*) for each trial exponential curve.
8. Plot the modelled data.



# Oxidative capacity analysis with {mnirs}

This article serves as a vignette for some of the kinetics analysis functions recently added to {mnirs}, and how they combine to perform robust OxCap analysis.

* `correct_blood_volume()` normalises values for oxy[haem], deoxy[haem], and total[haem] channels (if present), using an iterative method developed from @Beever2020 that properly accommodates negative values. See 
`?correct_blood_volume` for details.

* `extract_intervals()` detects specific events or times in an *"mnirs"* data frame, extracts an interval around each one, and returns a list of data frames The returned list of data frames is ready for further analysis. See 
`?extract_intervals` for details.

* `analyse_kinetics()` evaluates mNIRS response kinetics from a single *"mnirs"* data frame or a list of data frames, with a selection of parametric and non-parametric methods. It returns a formatted table of model results, with internal model components retrieved with `attributes(results)`. See 
`?analyse_kinetics` for details.\
\
We will use this function twice for OxCap analysis:

  * `analyse_kinetics(method = "peak_slope")` finds the peak linear regression slopes on deoxy[haem] from the extracted occlusion intervals. The steepest (positive or negative) slope is often interpreted as the point of greatest mismatch between O<sub>2</sub> supply and extraction. See `?peak_slope` for details on the vector-level function called by `analyse_kinetics()`.

  * `analyse_kinetics(method = "monoexponential")` then fits an exponential curve through the resulting slope values over time, to find the rate constant `k`. A self-starting function `SSmonoexponential()` is fit with `stats::nls()`, called as either a 4-parameter model with a time-delay parameter (TD), or a reduced 3-parameter model without TD where response onset is expected to be immediate. See `?monoexponential` and `?SSmonoexponential` for details.



## Setup

First, load our packages and initial setup options.

```{r}
#| label: setup
#| warning: false

library(dplyr)   ## for convenient data wrangling
library(ggplot2) ## for plotting
library(mnirs)   ## install: pak::pak("jemarnold/mnirs@dev")

options(mnirs.verbose = FALSE) ## silence mnirs messages
theme_set(theme_mnirs()) ## set ggplot2 theme
```


::: {.callout-note title="Acknowledgement"}
I received this example file from [Dr. Thomas Tripp](https://scholar.google.com/citations?user=A9JwQbYAAAAJ&hl=en), currently a postdoctoral fellow in [Dr. Martin MacInnis](https://scholar.google.com/citations?hl=en&user=tJepNIQAAAAJ)' lab at the University of Calgary. They have kindly allowed me to include this file with the {mnirs} package for users to examine themselves.

It can be accessed by calling `example_mnirs("portamon-oxcap")`.
:::



# Read {mnirs} data file

The included example file has three NIRS channels and a sample-number column, automatically converted to a time value by `read_mnirs()`. The file needs no other pre-processing steps.

Note: by default, *Artinis Oxysoft* exports an event label column without a named header — this is the column we will use to identify occlusion events by the label *"Occlusion"*. In {mnirs}, an unnamed column is coerced to name `"col_*"` with a numeric suffix for the column position in the data file. In our example file, this event column is therefore `"col_6"`. Not the most elegant solution, but it works for now.


```{r}
#| label: read file and process

df <- read_mnirs(
    file_path = example_mnirs("portamon-oxcap.xlsx"),
    ## identify and rename channels
    nirs_channels = c(thb = 2, hhb = 3, o2hb = 4),
    ## specify the event label column
    event_channel = c(event = "col_6")
)

## view the structure of our data frame
df
```

This file has two sequential occlusion trials. There are many ways to identify and separate trials. We will find the long gap between consecutive occlusion labels and split the data frame at that point.

```{r}
#| label: split df

## find large time gap between occlusion labels indicating trials
occl_times <- df$time[grepl("Occlusion", df$event)]
split_time <- occl_times[diff(occl_times) > 60] + 10

plot(df) +
    geom_vline(xintercept = split_time, linetype = "dashed")
```


The two exercise bouts and repeated occlusion trials are clearly identifiable from the raw data plot. A few resting occlusions were also performed at the start of the file, which we will ignore.



# Correct for blood volume

Before analysing the OxCap trials, the NIRS channels must be corrected for changes in blood volume during the repeated occlusions. This improves validity when evaluating NIRS slopes [@Ryan2012].

The preferred method proposed by @Beever2020 iteratively corrects for instantaneous changes in oxygenation between samples. `correct_blood_volume()` implements this method, further developed to accommodate negative NIRS values, which some mNIRS devices report.

## `correct_blood_volume()`

* `data`

    This function takes a data frame, corrects the specified NIRS channels for blood volume, and returns the data frame with the corrected `nirs_channels`.\
\
At least two of the three channels must be named explicitly as `oxy_channel`, `deoxy_channel`, and `total_channel`; the third is derived. These cannot be read from *"mnirs"* metadata, to avoid ambiguity. The corrected channels are passed on in metadata as `nirs_channels`.


```{r}
#| label: blood volume correction

df <- df |>
    correct_blood_volume(
        oxy_channel = o2hb,
        deoxy_channel = hhb,
        total_channel = thb
    )

## plot corrected data with occlusion event indicators
plot(df) +
    geom_vline(xintercept = occl_times, linetype = "dotted", alpha = 0.4)
```


This results in a symmetrical distribution of oxy- and deoxy[haem], with the sum total (total[haem]; THb) definitionally equal to zero at all samples.

Occlusions from the labels in `event_channel` are also plotted (vertical dotted lines) to visually confirm timing and detect any erroneous events.


# Extract occlusion intervals

The next step is to identify these occlusion events and extract each interval into a list of data frames for iterative analysis. We run this function twice to separate the occlusions within the two trials.


## `extract_intervals()`

* `data`

    This function takes a data frame or list of data frames, and returns a list of data frames, named per detected interval. *"mnirs"* metadata is passed to and from this function.

* `nirs_channels`, `time_channel`, `event_channel`, & `sample_rate`

    Specify which column names in `data` will be processed. `nirs_channels` are the response variables; `time_channel` is the predictor variable; `event_channel` specifies where to look for event labels; and `sample_rate` carries the measurement/exported samples-per-second rate used when ensemble-averaging across intervals. Unspecified arguments are retrieved from *"mnirs"* metadata. Channels not explicitly specified pass through unprocessed to the returned data frames.\
\
We will analyse OxCap from the deoxy[haem] (HHb) NIRS signal, as often recommended [@Costalat2025].

* `start` & `end`

    Interval boundaries are specified using helper functions: `by_time()` for `time_channel` values; `by_label()` for `event_channel` labels; `by_lap()` for lap indices in `event_channel`; or `by_sample()` for sample indices (row numbers). Provide both `start` and `end` to define precise intervals, or `start` alone with `span` to extract windows around events.\
\
First, `start` and `end` split the data frame between the two trials. Then, since the occlusions are all identified by the event label *"Occlusion"*, we detect this string in the `event` column as the `start` value to extract each interval.\
\
Detecting intervals by event labels must match a full string exactly or use a regular expression (regex) string.

* `group_intervals`

    Multiple events can be extracted for analysis as `"distinct"` intervals, or `"ensemble"`-averaged together. Different groupings can be ensemble-averaged by providing a list of intervals by number, e.g. `list(c(1, 2), c(3, 4))`.\
\
For manual grouping, interval numbers must be known ahead of time and always refer to order of occurrence in the data frame (sorted by `time_channel`). For this example, we analyse each interval separately using `group_intervals = "distinct"`.

* `span`

    A two-element vector of `time_channel` values specifying a time window `c(before, after)`, where positive values indicate time after the event and negative values indicate time before. When both `start` and `end` are provided, `span` shifts boundaries additively: `span[1]` adjusts starts, `span[2]` adjusts ends.\
\
Since all occlusion bouts were held for 5 s, we use a 5-sec time span. If occlusion duration changed through the protocol, we could specify a list of time spans (`span = list()`); here a single vector is recycled.\
\
Visual inspection (not shown) revealed that the deoxy[haem] signal was disrupted by cuff inflation during the first ~0.5 s of each occlusion interval, which might influence slope values. We therefore exclude the first second with `span = c(1, 5)`, i.e. include from second 1 (rather than 0) to 5 after each *"Occlusion"* label.

* `zero_time`

    Time values for each event interval can be recalculated to start from zero at the event indicator. Ensemble-averaged intervals are always re-calculated from zero, since original time values are redundant when ensembling.\
\
For repeated occlusion testing, the original time information per interval is needed to subsequently model the exponential recovery function, so we keep `zero_time = FALSE`.


```{r}
#| label: extract intervals
#| fig-width: 9
#| fig-asp: 0.8

## split the df at the identified `split_time` value
## trial 1 starts at time = 0, ends at `split_time`
## trial 2 starts at `split_time`, ends at the last time sample
df_list <- extract_intervals(
    df,
    group_intervals = "distinct",
    start = by_time(c(0, split_time)),
    end = by_time(split_time, tail(df$time, 1)),
    span = c(0, 0),   ## no buffer around the start/end boundaries
    zero_time = FALSE ## preserve original time values at each event
) |>
    ## pipe the list of two trial dfs into another `extract_intervals()`
    ## identify and extract intervals by event label "Occlusion"
    ## extract an interval from 1-5 sec after the label time
    extract_intervals(
        group_intervals = "distinct",
        start = by_label("Occlusion"),
        span = c(1, 5)
    )

## visualise all occlusion intervals for error checking
plot(df_list, ncol = 8) +
    theme(axis.text.x = element_blank())
```


As expected, on visual inspection the first interval in each repeated occlusion trial is the steepest, becoming progressively less steep at each subsequent occlusion as mV̇O<sub>2</sub> recovers.

Next, we calculate the slope value for each occlusion interval.


# Calculate occlusion slopes


## `analyse_kinetics()`

* `data`

    This function takes a data frame or list of data frames, analyses all channels within each separately, and returns a formatted table of results. Model components such as coefficients, fitted data frames, input metadata, and diagnostics can be retrieved from results metadata with `attributes()` or `names()`.

* `nirs_channels` & `time_channel`

    If the data contain *"mnirs"* metadata, these are detected automatically; otherwise they can be specified explicitly for each data frame, or as a single vector of channel names recycled across all data frames.

* `start_time`, `end_window` & `direction`

    These arguments apply to every method. `start_time` marks the response onset. When intervals come from `extract_intervals()`, it is retrieved automatically from *"mnirs"* metadata; otherwise it defaults to the first non-negative time value or zero.\
\
`end_window` defines the end of the fitting window by looking forward for a reasonable end time, taking the first peak or trough (for rising or falling trend, respectively) with no more extreme value within that span. The default `end_window = Inf` uses the global extreme across the whole interval, appropriate for these short occlusion windows.\
\
`direction` (default `"auto"`) detects whether the response rises or falls, or can be set explicitly to `"positive"` or `"negative"`. For parametric fits it also constrains the sign of the fitted amplitude `B − A`. For deoxy[haem], we specify `direction = "positive"`, since deoxygenation increases during occlusions.

* `method`

    Four analysis methods are currently available. This OxCap analysis needs two: *"peak_slope"* and *"monoexponential"*.



### Analyse peak slope kinetics

* `method = "peak_slope"`

The **peak slope** method is a *semi-parametric* approach: it fits a linear regression model across a sliding local window, returning the single steepest slope along with the time at which it occurs. For OxCap analysis, the steepest local rate of deoxygenation is interpreted as the point of greatest mismatch between O<sub>2</sub> delivery and extraction — a proxy for mV̇O<sub>2</sub> during each occlusion.

* `width`, `span`

    The local window is specified by either `width` (a number of samples) or `span` (a range of time in units of `time_channel`). Here we want the peak slope from a 3-sec span within each occlusion interval.

* `align`

    The window can be `"centre"`-aligned around the target sample (`idx`), `"left"`-aligned with `idx` at the start (forward-looking), or `"right"`-aligned with `idx` at the end (backward-looking). We keep the default `"centre"` alignment.

* `partial`

    By default, a local window only returns a linear regression model if all samples within it are valid. If `partial = TRUE`, windows return a slope model as long as two or more samples are available.\
\
Windows with fewer samples tend to return steeper slopes, which may not be relevant to our interpretation — e.g. at the edges of a noisy dataset, two samples alone can produce an extreme local slope that we would usually want to ignore.


```{r}
#| label: find peak slopes
#| warning: false
#| fig-width: 9
#| fig-asp: 0.8

## find peak 3-sec hhb slopes within each occlusion interval
slope_results <- analyse_kinetics(
    df_list,
    nirs_channels = hhb,
    time_channel = time,
    method = "peak_slope",
    span = 3,
    direction = "positive",
    partial = FALSE,
)

## view the formatted results
slope_results

plot(slope_results, ncol = 8, label_size = 3.5) +
    theme(axis.text.x = element_blank())
```

Individual kinetics results can be quickly visualised with the built-in `plot()` function (see `?plot.mnirs_kinetics`), but we will build a custom plot below to better visualise the exponential response kinetics for mV̇O<sub>2</sub> in each trial.

Next, we extract the slope values into their own data frame for the kinetics analysis.


```{r}
#| label: create slopes_df

## extract interval start times from kinetics metadata
start_times <- slope_results$interval_times$start_times

## extract a data frame with slope coef & time values
## grouped by trial, identified by the interval names
## `peak_slope_time` is a delta from each interval `start_time`
slopes_df <- slope_results$coefficients |>
    mutate(
        trial = if_else(grepl("interval_1", interval), 1, 2),
        trial = factor(trial),
        time = start_times + peak_slope_time
    ) |>
    group_by(trial) |>
    ## recalculate time starting from zero for each trial
    mutate(time = time - time[1L]) |>
    select(trial, time, slope) |> 
    ## add mnirs metadata
    create_mnirs_data(
        nirs_channels = slope,
        time_channel = time
    )
```


```{r}
#| label: plot peak slopes

## plot observed slope values for each trial
p <- ggplot(slopes_df, aes(x = time, y = slope, colour = trial)) +
    coord_cartesian(ylim = c(0, NA)) +
    scale_x_continuous(
        name = "Time (mm:ss)",
        breaks = breaks_timespan(),
        labels = format_hmmss
    ) +
    scale_y_continuous(
        name = expression(bold(HHb~Slope~'(μM'%.%sec^'-1'*')')),
    ) +
    scale_colour_mnirs() +
    geom_line() +
    geom_point(fill = "white", size = 3, shape = 21, stroke = 1)

p
```



# Model mV̇O<sub>2</sub> recovery


## `analyse_kinetics()`


### Analyse monoexponential kinetics

* `method = "monoexponential"`

The **monoexponential** method is a *parametric* approach: it fits a self-starting exponential curve to time series data with `nls()`, using `SSmonoexponential()` to initialise. This models how mV̇O<sub>2</sub> recovers back to baseline after the exercise stimulus, and its rate constant *k* is the OxCap outcome.

It fits either a **4-parameter** model `A + (B − A)(1 − exp(−(t − TD) / tau))` with a time delay *TD*, or a reduced **3-parameter** model without *TD* when the response is expected to start immediately.

* `use_TD` (*TD* coefficient)

    An argument specific to *"monoexponential"* method: it explicitly specifies a 4-parameter model with a time delay between the stimulus and the start of a systematic exponential response. The model falls back to a 3-parameter fit if the 4-parameter fit fails to converge.\
\
During repeated occlusions after exercise, the first occlusions may not conform to an exponential response, i.e. if oxygen saturation is too low and limiting to mV̇O<sub>2</sub>. A time delay parameter can capture the true monoexponential response in these cases. This protocol was designed to avoid that delay, so *TD* was not required.

* *A*, *B* coefficients

    The starting (baseline) and ending (asymptote) values of the monoexponential curve, defining either exponential association (positive direction, *B* > *A*) or decay (negative direction, *A* > *B*).

* *tau* coefficient

    The time constant (τ) describes the speed of the exponential response from *A* to *B*, approximated by the time elapsed at 63.2% of the total response amplitude — i.e. when predicted `y` equals `A + 0.632 × (B − A)`.\
\
The rate constant (*k*) is the inverse of the time constant (`k = 1 / tau`) and is also commonly used in NIRS research. *k* is returned as the inverse of `time_channel` units (seconds) by default; here we convert *k* to min<sup>-1</sup>.


```{r}
#| label: monoexponential curve modelling

## use 3-param monoexponential model without TD
## fit models grouped by trial
## channels passed via mnirs metadata added above
monoexp_results <- analyse_kinetics(
    slopes_df,
    method = "monoexponential",
    use_TD = FALSE
)

## view the formatted results
monoexp_results
```

```{r}
#| label: build coef_labels

## extract and bind the list of fitted data for plotting
fitted_data <- do.call(rbind, monoexp_results$data)

## extract tau and k coef labels per trial, for plotting
coef_labels <- monoexp_results$coefficients |>
    rename(trial = interval) |>
    summarise(
        .by = trial,
        label = sprintf(
            "Trial %s\ntau = %s sec\nk = %s min⁻¹",
            trial,
            round(tau, 1),
            round(k * 60, 1) ## convert to min^-1
        )
    )
```

```{r}
#| label: plot modelled data

## add fitted kinetics curves to the plot of observed data
p +
    geom_line(
        data = fitted_data,
        aes(y = slope_fitted, group = trial),
        linewidth = 0.8, colour = "black"
    ) +
    geom_text(
        data = coef_labels, show.legend = FALSE,
        aes(label = label, colour = trial, group = trial),
        x = Inf, y = Inf,
        size = 5, hjust = 1.1, vjust = c(1.5, 3)
    )
```

# Conclusion

This article demonstrates how muscle oxidative capacity analysis can be performed using {mnirs} with relatively simple, reproducible data wrangling steps. Plotting incremental steps along the way adds some complexity to the code, but adds valuable visual validation which I encourage whenever performing multi-stage analyses.

Future development could wrap these steps into a dedicated convenience function with recommended default parameters, streamlining the analysis process. However, there are always trade-offs between automation and customisation: fitting nonlinear models to sparse, noisy biological data can produce unstable fit parameters or fail to converge. Manual processing steps may still be needed to accommodate these limitations.

# References