---
title: "Reading PIONIRS data with mnirs"
author:
    - name: "Jem Arnold  [{{< fa brands github >}}](https://github.com/jemarnold/mnirs){target=\"_blank\"} [{{< fa brands bluesky >}}](https://bsky.app/profile/jemarnold.bsky.social){target=\"_blank\"} [{{< ai researchgate >}}](https://researchgate.net/profile/Jem-Arnold){target=\"_blank\"} [{{< ai orcid >}}](https://orcid.org/0000-0003-3908-9447){target=\"_blank\"}"
date: last-modified
format:
    html:
        embed-resources: true
        toc: true
csl: ../apa.csl
knitr:
    opts_chunk:
        collapse: true
        comment: '#>'
        dpi: 300
        out.width: "100%"
        fig.width: 5
        fig.asp: 0.667
        fig.align: "center"
---

# Introduction

*{mnirs}* can now read *PIONIRS NIRSBOX* `.ftn` and `.ftn2` file exports directly with `read_mnirs()`. This article demonstrates reading both file types and running a short processing and analysis pipeline on an arterial occlusion recording.

::: {.callout-note}
This article assumes basic familiarity with the *{mnirs}* package. See [*Reading and Cleaning Data with mnirs*](https://jemarnold.github.io/mnirs/articles/reading-mnirs-data.html){target="_blank"} for an overview.
:::


# Setup

```{r}
#| label: setup
#| warning: false

library(ggplot2) ## for plotting
library(mnirs)   ## install: pak::pak("jemarnold/mnirs@dev")

```

```{r}
#| label: file paths
#| echo: false
PIONIRS_ftn <- r"(C:\R-Projects\mnirs\tests\testthat\testdata\pionirs_occlusion.ftn)"
PIONIRS_ftn2 <- r"(C:\R-Projects\mnirs\tests\testthat\testdata\pionirs_occlusion.ftn2)"
```


# Automatic `.ftn(2)` file channel detection

*PIONIRS* `.ftn(2)` files are recognised automatically. With no channels specified, `read_mnirs()` returns `"Time"` as `time_channel`, `"TagLabel"` as `event_channel`, and any *"StO2"* channels as `nirs_channels`.

When automatically detecting a known NIRS device file format, all columns are returned to allow exploration of the file.

```{r}
#| label: read ftn2

## read an example file with no additional parameters
## file paths are hidden for this example
df_sto2 <- read_mnirs(PIONIRS_ftn2)

print(df_sto2, n = 5)
```


## Filtering the data

`filter_mnirs()` applies a digital filter to all `nirs_channels` automatically retrieved from metadata in the data frame of class *"mnirs"* read above.

Here, a Butterworth low-pass 2nd-order filter with cutoff frequency 0.05 Hz seems to smooth the data sufficiently.

```{r}
#| label: filter sto2

df_filt <- filter_mnirs(
    df_sto2,
    method = "butterworth",
    order  = 2,
    fc     = 0.05
)

## overlay unfiltered data for comparison
plot(df_filt) +
    geom_line(
        data  = df_sto2,
        aes(y = `StO2(CH1)`, colour = "StO2(CH1)"),
        alpha = 0.3
    ) +
    geom_line(
        data  = df_sto2,
        aes(y = `StO2(CH2)`, colour = "StO2(CH2)"),
        alpha = 0.3
    )
```


# Specify explicit NIRS channels for analysis

Channels can also be specified and renamed explicitly. Here we read the raw haemoglobin concentration channels from the single-channel `.ftn` file, with the event column containing manually entered labels.

```{r}
#| label: read ftn

df_raw <- read_mnirs(
    file_path     = PIONIRS_ftn,
    nirs_channels = c(o2hb = "O2Hb", hhb = "HHb", thb = "THb"),
    time_channel  = c(time = "Time"),
    event_channel = c(labels = "TagLabel"),
    verbose = FALSE
) |> 
    filter_mnirs(method = "butterworth", order = 2, fc = 0.05)

plot(df_raw) + 
    geom_vline(
        data = subset(df_raw, !is.na(labels)),
        aes(xintercept = time), 
        linetype = "dotted"
    )
```


## Correct for blood volume: Δtotal[haem]

`correct_blood_volume()` normalises oxy-, deoxy-, and total[haem] for changes in blood volume during the occlusion. Channels must be specified explicitly to avoid naming ambiguity.

```{r}
#| label: blood volume correction

df_corr <- correct_blood_volume(
    df_raw,
    oxy_channel   = o2hb,
    deoxy_channel = hhb,
    total_channel = thb
)

plot(df_corr)
```


## Extract deoxygenation & reoxygenation intervals

`extract_intervals()` can locate the labels *"Occlusion start"* and *"Recovery"*. an `end` value can also be specified, or a timespan (in units of `time_channel`) can be defined around each detected event label, to return the desired interval lengths.

O<sub>2</sub> extraction or rate of *mV̇O<sub>2</sub>* can be estimated from the peak 120-sec linear slope during the occlusion deoxygenation (*"slope 1"*).

Reoxygenation kinetics or *microvascular responsiveness* is typically estimated as the peak 10-sec slope after occlusion (*"slope 2"*), or with a monoexponential model. We will extract the full reoxygenation and hyperaemia phase for analysis.

We will isolate the *"deoxy[haem]"* channel to analyse, since *"hhb"* and *"o2hb"* are now mirror images of each other.

```{r}
#| label: extract intervals

df_list <- extract_intervals(
    df_corr,
    nirs_channels = hhb,
    start = by_label("Occlusion", "Recovery"),
    span  = c(0, 300)
)

print(df_list, n = 5)
```


# Analyse microvascular responsiveness

## Peak linear regression slope

`analyse_kinetics(method = "peak_slope")` fits a rolling linear regression algorithm across the full interval, and returns the greatest (positive or negative, depending on direction) slope centred within the specified time `span`.

We can specify different `span` parameter for each interval with a named `list()`.

```{r}
#| label: peak_slope

result_deoxy <- analyse_kinetics(
    df_list,
    method = "peak_slope",
    span   = list(interval_1 = 120, interval_2 = 10)
) |>
    print()

plot(result_deoxy)
```


# Analyse reoxygenation kinetics

Now let's look at some of the more advanced kinetics analysis methods we can fit during the reoxygenation and hyperaemia window.

## Monoexponential

`method = "monoexponential"` fits an exponential curve to the data. `end_window = 30` dynamically ends the fitting window at the first peak or trough (depending on direction) with no more extreme values within 30 sec, avoiding the post-hyperaemic decay toward baseline which is non-asymptotic and would bias the monoexponential time constant coefficient.

```{r}
#| label: monoexponential

result_monoexp <- analyse_kinetics(
    df_list[2],
    method     = "monoexponential",
    end_window = 30  ## includes 30-sec after the first local extrema
) |>
    print()

plot(result_monoexp, time_labels = TRUE)
```

With experimental kinetics analysis methods (still under development and not yet validated), we can avoid this potential for coefficient bias by modelling the hyperaemic drift as a secondary linear slope component in a two-phase *"exponential-linear"* kinetics response.


## Exponential-drift

`method = "exponential_drift"` (or `"exponential_linear"`) fits a linear slope term after the primary monoexponential response, to separate the fast reoxygenation phase from the slow hyperaemic drift.

```{r}
#| label: exponential drift

result_drift <- analyse_kinetics(
    df_list[2],
    method     = "exponential_drift",
    end_window = 180  ## includes more of the hyperaemic response
) |>
    print()

plot(result_drift, time_labels = TRUE)
```


## Biexponential

`method = "biexponential"` is another experimental method which fits a two-phase response with overlapping fast and slow monoexponential curves.

```{r}
#| label: biexponential

result_biexp <- analyse_kinetics(
    df_list[2],
    method = "biexponential"
    ## `end_window = Inf` by default; fitting to the full data available
) |>
    print()

plot(result_biexp, time_labels = TRUE)
```


## Compare fit diagnostics

Fit diagnostics are stored in `results$diagnostics` and can be used to statistically and qualitatively compare between models.

Information criteria (`aic`, `aicc`, `bic`) can be used to compare between models with different numbers of parameters (`n_params`). Lower values represent statistically "better" fitting models, after penalising extra parameters (each criterion uses slightly different weights and may disagree on marginal differences).

However, in this case, because each model is fit on a different number of samples (`n_obs`) in addition to parameters, the information criteria *cannot* be used to statistically conclude better fit.

The other model diagnostics can be used as qualitative comparators, and the "best" model can be chosen based on physiological expectations, experimental design, and established validity.

::: {.callout-caution}
Two-phase **Exponential-drift** and **Biexponential** kinetics are currently experimental methods and have not yet been validated.
:::

```{r}
#| label: compare diagnostics

rbind(
    monoexponential   = result_monoexp$diagnostics,
    exponential_drift = result_drift$diagnostics,
    biexponential     = result_biexp$diagnostics
)
```

In this example, `r2`, `rmse`, and `snr` (signal-to-noise ratio) suggest that the biexponential model is not capturing the shape the full hyperaemic decay and return to baseline, as well as the exponential-linear drift model over the hyperaemic component alone.


# Acknowledgements

Thanks to *Marianna Neri*, [*Dr. Simone Porcelli*](https://medicinamolecolare.dip.unipv.it/en/research/research-teams-and-topics/human-physiology/human-integrative-physology-exercise-simone){target="_blank"}, and the developers at [*PIONIRS*](https://www.pionirs.com/wp/){target="_blank"} for providing access to these example files to allow me to integrate this excellent device into *{mnirs}*.