---
title: "mnirs — Bluesky posts"
format:
  html:
    self-contained: true
    code-tools: false
    include-in-header:
      text: |
        <style>
          main.content {
            max-width: 680px;
            margin-left: auto;
            margin-right: auto;
          }
        </style>
execute:
  echo: true
  message: false
  warning: false
  fig-dpi: 600
  cache: true
---

```{r setup}
#| include: false
library(mnirs)
library(ggplot2)
options(mnirs.verbose = FALSE)
## print 5 rows only
options(pillar.print_min = 5, pillar.print_max = 5)
```

## `read_mnirs()`

Use #mnirs 🦵🔦 to easily read files and plot data from wearable NIRS devices

Automatically detect NIRS channels from @MoxyMonitor @TrainRedSensors @Artinis_MS and others. or specify and rename custom channels

\#rstats

```{r read_mnirs}
library(mnirs)

## automatically detect or specify and rename channels
data_mnirs <- read_mnirs(
    file_path      = example_mnirs("moxy_ramp"),
    nirs_channels  = c(
        smo2_left  = "SmO2 Live",
        smo2_right = "SmO2 Live(2)"
    ),
    time_channel   = c(time = "hh:mm:ss"),
    event_channel  = c(lap = "Lap")
)

## built-in plotting with `{ggplot2}`
plot(data_mnirs)
```

---

## `replace_mnirs()`

This is how #mnirs 🦵🔦 can easily replace data spikes, signal dropouts, local outliers, and invalid data

\#rstats

```{r replace_mnirs_display}
#| eval: false

library(mnirs)

data_cleaned <- replace_mnirs(
    data           = data_mnirs, # data of class "mnirs"
    invalid_values = 0,          # specify invalid values
    invalid_above  = 90,         # specify invalid_above/below range
    outlier_cutoff = 3,          # median deviation (± SD) outlier filter
    span           = 15,         # local outlier window in `time_channel` units
    method         = "linear"    # linear interpolation over missing/replaced
)

plot(data_cleaned, time_labels = TRUE)
```

```{r replace_mnirs_run}
#| echo: false
#| eval: true

data_cleaned <- replace_mnirs(
    data_mnirs,
    invalid_values = 0,
    invalid_above = 90,
    outlier_cutoff = 3,
    span = 15,
    method = "linear"
)

which_diff <- which(abs(data_mnirs$smo2_left - data_cleaned$smo2_left) >= 5)

plot(data_cleaned, time_labels = TRUE) +
    scale_y_continuous(expand = expansion(0.03)) +
    geom_line(
        data = data_mnirs,
        aes(y = smo2_left, colour = "smo2_left"),
        alpha = 0.3
    ) +
    geom_point(
        data = data_mnirs[which_diff, ],
        aes(y = smo2_left),
        colour = palette_mnirs("red"),
        shape = 4,
        size = 5
    )
```



---

## `filter_mnirs()`

Use #mnirs 🦵🔦 to easily smooth noisy NIRS data

We can compare different digital filter methods with custom parameters

Simple *"moving average"* or *"Butterworth"* low-pass filters are most common

\#rstats 

```{r filter_mnirs_display}
#| eval: false

library(mnirs)

data_filtered <- filter_mnirs(
    data_cleaned,           # data of class "mnirs" 
    method = "butterworth", # specify a filter method
    order  = 2,             # filter order (number of passes)
    fc     = 0.02,          # cutoff frequency (Hz or fractional)
    type   = "low",         # specify filter type
    na.rm  = TRUE           # ignore missing data (`NA`)
)

plot(data_filtered, time_labels = TRUE)
```

```{r filter_mnirs_run}
#| echo: false
#| eval: true

data_filtered <- filter_mnirs(
    data_cleaned,
    method = "butterworth",
    order = 2,
    fc = 0.02,
    type = "low",
    na.rm = TRUE
)

plot(data_filtered, time_labels = TRUE) +
    coord_cartesian(ylim = c(0, 100)) +
    geom_line(
        data = data_cleaned,
        aes(y = smo2_left, colour = "smo2_left"),
        alpha = 0.3
    ) +
    geom_line(
        data = data_cleaned,
        aes(y = smo2_right, colour = "smo2_right"),
        alpha = 0.3
    )
```


---

## `rescale_mnirs()`

Use #mnirs 🦵🔦 to easily rescale NIRS signal range

Process channels independently to a functional 0-100% exercise range like this ramp test, or a physiological range with occlusions

\#rstats

```{r rescale_mnirs_display}
#| eval: false

library(mnirs)

data_rescaled <- rescale_mnirs(
    data_filtered,               # data of class "mnirs"  
    group_channels = "distinct", # process channels independently
    range          = c(0, 100)   # define the new signal range
)

plot(data_rescaled, time_labels = TRUE)
```

```{r rescale_mnirs_run}
#| echo: false
#| eval: true

data_rescaled <- rescale_mnirs(
    data_filtered,
    group_channels = "distinct",
    range = c(0, 100)
)

plot(data_rescaled, time_labels = TRUE) +
    geom_hline(yintercept = c(0, 100), linetype = "dotted")
```

---

Use #mnirs 🦵🔦 to easily rescale NIRS signal range

Process channels together to preserve relative scale between muscle sites, or for absolute NIRS signals (oxy/O<sub>2</sub>Hb, deoxy/HHb)

\#rstats 

```{r rescale_mnirs_display2}
#| eval: false

library(mnirs)

data_rescaled <- rescale_mnirs(
    data_filtered,               # data of class "mnirs"
    group_channels = "ensemble", # process channels together
    range          = c(0, 100)   # define the new signal range
)

plot(data_rescaled, time_labels = TRUE)
```

```{r rescale_mnirs_run2}
#| echo: false
#| eval: true

data_rescaled <- rescale_mnirs(
    data_filtered,
    group_channels = "ensemble",
    range = c(0, 100)
)

plot(data_rescaled, time_labels = TRUE) +
    geom_hline(yintercept = c(0, 100), linetype = "dotted")
```

---

## `shift_mnirs()`

Use #mnirs 🦵🔦 to easily shift NIRS signal range

Signals can be transformed to a change from baseline while preserving relative scale

\#rstats

```{r shift_mnirs_display}
#| eval: false

library(mnirs)

data_shifted <- shift_mnirs(
    data_filtered,               # data of class "mnirs"
    group_channels = "distinct", # process channels separately
    to             = 0,          # shift the target to zero
    span           = 120,        # target a 2-min mean
    position       = "first"     # at the starting baseline
)

plot(data_shifted, time_labels = TRUE)
```

```{r shift_mnirs_run}
#| echo: false
#| eval: true

data_shifted <- shift_mnirs(
    data_filtered,                              
    group_channels = "distinct",
    to             = 0,                          
    span           = 120,                        
    position       = "first"                     
)

plot(data_shifted, time_labels = TRUE) +
    geom_hline(yintercept = 0, linetype = "dotted") +
    geom_vline(xintercept = 120, linetype = "dotted")
```

---


## `extract_intervals()`

Use #mnirs 🦵🔦 to easily extract intervals from laps, times, or event labels

Extract intervals independently as a list, or ensemble-average to a single data frame

\#rstats

```{r load_ensemble_data}
#| echo: false

data_intervals <- read_mnirs(
    file_path = example_mnirs("moxy_interval"),
    nirs_channels = c(smo2 = "SmO2 Live"),
    time_channel = c(time = "hh:mm:ss"),
    event_channel = c(lap = "Lap"),
)
```

```{r extract_intervals_distinct}
extract_intervals(
    data_intervals,
    start = by_time(124, 486, 848), # specify times in `time_channel` units
    end = by_lap(3, 5, 7),          # or `lap` integer values
    span = c(-60, 0),               # expand bounds around `c(start, end)`
    group_intervals = "distinct",   # return each interval in a list
) |>
    ## plot lists of class "mnirs" as facets
    plot()
```

```{r extract_intervals_ensemble}
extract_intervals(
    data_intervals,
    start = by_time(124, 486, 848),
    end = by_lap(3, 5, 7),
    span = c(-60, 0),
    group_intervals = "ensemble", # ensemble-average intervals
) |> 
    plot(time_labels = TRUE)
```


```{r extract_intervals_group}

data_ensembled <- extract_intervals(
    data_intervals,
    start = by_lap(2, 4, 6, 3, 5, 7),
    end = by_lap(2, 4, 6, 3, 5, 7),
    span = list(c(-30, -60), c(-30, 0)),
    group_intervals = list(c(1, 2, 3), c(4, 5, 6)),
    zero_time = TRUE
)

plot(data_ensembled)
```

## `analyse_kinetics()`

Use #mnirs 🦵🔦 to easily analyse deoxygenation & reoxygenation kinetics

```{r analyse_kinetics}

analyse_kinetics(
    data_ensembled,
    method = "biexp",
    # start_time = c(),
    verbose = TRUE
) |> 
    plot()
```
