---
title: "mnirs 0.7.0 — social posts"
format:
    html:
        embed-resources: true
        toc: false
execute:
  echo: true
  message: false
  warning: false
  dpi: 300
  fig-align: center
  out-width: "100%"
knitr:
    opts_chunk:
        eval: false
        collapse: true
        comment: '#>'
---

## Post 1 — Intro

{mnirs} R package for reading, processing, and analysing muscle near-infrared spectroscopy data🦵🔦 updated to 0.7.0 🎉

Mostly internal changes getting ready for `mnirs::analyse_kinetics()` (soon!)

Here are the key changes: 1/5🧵🔗👇 #rstats #nirs #muscleoxygenation


![](../../man/figures/mnirs-hex.svg){width=200}

------------------------------------------------------------------------

## Post 2 — Lists & grouped data frames

Core processing functions now accept a list of data frames or grouped dfs with `dplyr::group_by`

You could `filter_mnirs()` over a list of dfs from multiple `read_mnirs()` files,

Or `extract_intervals()` from a longer session and `rescale_mnirs()` each independently

```{r}
#| eval: true
#| echo: false

library(mnirs)
library(ggplot2)

options(pillar.print_max = 5, pillar.print_min = 5)

theme_set(theme_mnirs())

data <- read_mnirs(
    file_path = example_mnirs("moxy_interval"),
    nirs_channels = c(
        smo2 = "SmO2 Live",
        thb = "THb"
    ),
    time_channel = c(time = "hh:mm:ss")
)

data_list <- extract_intervals(
    data,
    nirs_channels = smo2,
    group_intervals = "distinct",
    start = by_time(124, 1210),
    end = by_time(364, 1450),
    span = c(-60, 120),
    zero_time = TRUE
)

df1 <- data_list[[1L]]
df2 <- data_list[[2L]]
df1$thb <- NULL
df2$thb <- NULL
```

```{r}
#| eval: true

data_list <- filter_mnirs(
    data = list(df1, df2),
    method = "butterworth",
    type = "low",
    order = 2,
    W = 0.2
)

data_list

plot(data_list)
```

------------------------------------------------------------------------

## Post 3 — Per-channel arguments

{mnirs} 0.7.0 can now process each nirs_channel with independent settings:

You can replace invalid values with different criteria for "smo2" and "thb"

Or filter_mnirs() one channel with "moving average" and another with "butterworth", side by side.

```{r}
#| eval: true
#| echo: false



lo <- c(275, 319, 490, 496, 550, 700, 702, 710, 722)
hi <- c(240, 244, 800)
data$smo2[lo] <- 0
data$smo2[hi] <- 100
set.seed(12)
data$thb[lo] <- data$thb[lo] + rnorm(length(lo), 0, 1)
data$thb[hi] <- data$thb[hi] + rnorm(length(hi), 0, 1)

data <- rescale_mnirs(data, group_channels = "distinct", range = list(smo2 = c(0, 100), thb = c(00, 20)))

```

```{r}
#| eval: true
#| echo: true

data_clean <- replace_mnirs(
    data,
    invalid_above = list(smo2 = 80, thb = 13), ## assign independent values
    invalid_below = list(thb = 10),  ## assign only one channel, ...
    invalid_values = list(smo2 = 0), ## or the other (with a warning)
    method = "none"                  ## replace with NA
)

## rows where invalid values were replaced
invalid <- data[is.na(data_clean$smo2) | is.na(data_clean$thb), ]

## plot the clean data and overlay replaced invalid values
plot(data_clean, time_labels = TRUE) +
    geom_line(data = data, aes(y = smo2, colour = "smo2"), alpha = 0.2) +
    geom_point(data = invalid, aes(y = smo2, colour = "smo2")) +

    geom_line(data = data, aes(y = thb, colour = "thb"), alpha = 0.2) +
    geom_point(data = invalid, aes(y = thb, colour = "thb"))
```


------------------------------------------------------------------------

## Post 4 — `group_channels`

A minor breaking change ⚠️ to `shift_mnirs()`, `replace_mnirs()`, and `extract_mnirs()`

`group_channels` arg handles channels

`group_intervals` arg handles intervals for `extract_intervals()` 

Options are "distinct", "ensemble", or a custom group list

```{r}
#| eval: true

data_rescaled <- rescale_mnirs(
    data_clean,
    nirs_channels = c(smo2, thb), ## pick the channels
    group_channels = "distinct",  ## how they are grouped
    range = c(0, 100),            ## same range for both channels
)

## `group_intervals replaces `event_groups`
ensemble <- extract_intervals(
    data_rescaled,
    group_intervals = "ensemble",
    start = by_time(124, 486, 848, 1210), ## extract four intervals,
    end = by_time(364, 726, 1088, 1450),  ## return one ensemble-averaged
    span = c(-10, 60)
)

ensemble

plot(ensemble, time_labels = TRUE)
```

------------------------------------------------------------------------

## Post 5 — link

{mnirs} 0.7.0 can be downloaded from CRAN 🧑‍💻: `install.packages("mnirs")`

More information on the {mnirs} package website: https://jemarnold.github.io/mnirs/index.html

