---
title: "mnirs — Bluesky posts"
format:
  html:
    self-contained: true
    code-tools: false
    include-in-header:
      text: |
        <style>
          main.content {
            max-width: 680px;
            margin-left: auto;
            margin-right: auto;
          }
        </style>
execute:
  echo: true
  message: false
  warning: false
  fig-dpi: 600
  cache: true
---

```{r setup}
#| include: false
library(mnirs)
library(ggplot2)
library(patchwork)
options(mnirs.verbose = FALSE)
```

## `read_mnirs()`

With {mnirs} we can easily read files and plot data from mNIRS devices

Example files are included from @MoxyMonitor @TrainRedSensors & @Artinis_MS 

But what about cleaning those data spikes? 🤔

\#rstats #mnirs 🦵🔦

```{r read_mnirs}
library(mnirs)

data_mnirs <- read_mnirs(
    file_path      = example_mnirs("moxy_ramp"), # file path
    nirs_channels  = c(
        smo2_left  = "SmO2 Live",         # specify and rename NIRS columns
        smo2_right = "SmO2 Live(2)"
    ),
    time_channel   = c(time = "hh:mm:ss") # specify time/sample column
)

plot(data_mnirs)
```

---

## `replace_mnirs()`

This is how we can easily replace data spikes, dropouts, outliers, and invalid values with {mnirs}

But what about smoothing/filtering the data? 🤔

\#rstats #mnirs 🦵🔦

```{r replace_mnirs_display}
#| eval: false

library(mnirs)

data_cleaned <- replace_mnirs(
    data_mnirs,                 # data of class `"mnirs"`
    invalid_values = c(0, 100), # specify invalid values
    outlier_cutoff = 3,         # median abs deviation (± SD) outlier filter
    span           = 15,        # local outlier window in `time_channel` units
    method         = "linear"   # linear interpolation over missing/replaced
)

plot(data_cleaned, time_labels = TRUE)
```

```{r replace_mnirs_run}
#| echo: false
#| eval: true

data_cleaned <- replace_mnirs(
    data_mnirs,
    invalid_values = c(0, 100),
    outlier_cutoff = 3,
    span = 7,
    method = "linear"
)

which_diff <- which(abs(data_mnirs$smo2_left - data_cleaned$smo2_left) >= 5)

plot(data_cleaned, time_labels = TRUE) +
    geom_line(
        data = data_mnirs,
        aes(y = smo2_left, colour = "smo2_left"),
        alpha = 0.3
    ) +
    geom_point(
        data = data_mnirs[which_diff, ],
        aes(y = smo2_left),
        colour = palette_mnirs("red"),
        shape = 4,
        size = 5
    )
```

---

## `filter_mnirs()`

We can easily smooth our noisy mNIRS data with {mnirs}

This is how to apply a low-pass Butterworth filter with custom parameters

But what if I want to rescale the data? 🤔

\#rstats #mnirs 🦵🔦

```{r filter_mnirs_display}
#| eval: false

library(mnirs)

data_filtered <- filter_mnirs(
    data_cleaned,           # data of class `"mnirs"` 
    method = "butterworth", # specify a digital filter method
    order  = 2,             # filter order (number of passes)
    fc     = 0.02,          # cutoff frequency (Hz)
    type   = "low",         # specify filter type
    na.rm  = TRUE           # ignore missing data (`NA`)
)

plot(data_filtered, time_labels = TRUE)
```

```{r filter_mnirs_run}
#| echo: false
#| eval: true

data_filtered <- filter_mnirs(
    data_cleaned,
    method = "butterworth",
    order = 2,
    fc = 0.02,
    type = "low",
    na.rm = TRUE
)

plot(data_filtered, time_labels = TRUE) +
    geom_line(
        data = data_cleaned,
        aes(y = smo2_left, colour = "smo2_left"),
        alpha = 0.3
    ) +
    geom_line(
        data = data_cleaned,
        aes(y = smo2_right, colour = "smo2_right"),
        alpha = 0.3
    )
```

---

## `rescale_mnirs()`

We can rescale our mNIRS data to a new functional range with {mnirs}

This is how to rescale each nirs channel to 0-100%

But how can I preserve relative scaling between nirs channels? 🤔

\#rstats #mnirs 🦵🔦

```{r rescale_mnirs_display}
#| eval: false

library(mnirs)

data_rescaled <- rescale_mnirs(
    data_filtered,                               # data of class `"mnirs"`  
    nirs_channels = list(smo2_left, smo2_right), # channels grouped separately
    range         = c(0, 100)                    # define the new dynamic range
)

plot(data_rescaled, time_labels = TRUE)
```

```{r rescale_mnirs_run}
#| echo: false
#| eval: true

data_rescaled <- rescale_mnirs(
    data_filtered,
    nirs_channels = list(smo2_left, smo2_right),
    range = c(0, 100)
)

plot(data_rescaled, time_labels = TRUE) +
    geom_hline(yintercept = c(0, 100), linetype = "dotted")
```

---

We can rescale our mNIRS data to a new functional range with {mnirs}

This is how to change the dynamic range but preserve relative scale between nirs channels

But what if we want to shift the baseline value? 🤔

\#rstats #mnirs 🦵🔦

```{r rescale_mnirs_display2}
#| eval: false

library(mnirs)

data_rescaled <- rescale_mnirs(
    data_filtered,                            # data of class `"mnirs"`  
    nirs_channels = c(smo2_left, smo2_right), # channels grouped together
    range         = c(0, 100)                 # define the new dynamic range
)

plot(data_rescaled, time_labels = TRUE)
```

```{r rescale_mnirs_run2}
#| echo: false
#| eval: true

data_rescaled <- rescale_mnirs(
    data_filtered,
    nirs_channels = c(smo2_left, smo2_right),
    range = c(0, 100)
)

plot(data_rescaled, time_labels = TRUE) +
    geom_hline(yintercept = c(0, 100), linetype = "dotted")
```

---

## `shift_mnirs()`

This is how we can shift our baseline value to zero with {mnirs}

mNIRS can be transformed to a change from baseline

But what if I want to extract the work and recovery intervals separately? 🤔

\#rstats #mnirs 🦵🔦

```{r shift_mnirs_display}
#| eval: false

library(mnirs)

data_shifted <- shift_mnirs(
    data_filtered,                               # data of class `"mnirs"`  
    nirs_channels = list(smo2_left, smo2_right), # channels grouped separately
    to            = 0,                           # shift the target to zero
    span          = 120,                         # target a 2-min mean
    position      = "first"                      # at the starting baseline
)

plot(data_shifted, time_labels = TRUE)
```

```{r shift_mnirs_run}
#| echo: false
#| eval: true

data_shifted <- shift_mnirs(
    data_filtered,                              
    nirs_channels = list(smo2_left, smo2_right),
    to            = 0,                          
    span          = 120,                        
    position      = "first"                     
)

plot(data_shifted, time_labels = TRUE) +
    geom_hline(yintercept = 0, linetype = "dotted")
```

---


## `extract_intervals()`

We can easily extract intervals with {mnirs}

This is how we can specify the interval start and end by time values

\#rstats #mnirs 🦵🔦

```{r extract_intervals}

interval_list <- mnirs::extract_intervals(
    data_filtered,
    start        = by_time(204, 865),  # specify times in `time_channel` units
    end          = by_time(865, 1050),
    span         = c(0, 0),            # expand boundaries around `c(start, end)`
    event_groups = "distinct",         # return each interval in a list
    zero_time    = TRUE                # rebase interval start times to zero
)

plot(interval_list[[1]]) + plot(interval_list[[2]])

```
